### 2016-08-28 previous-paper-analysis/summary.tex
Added previous paper analysis boilerplate

\documentclass[12pt,letterpaper]{article}
\usepackage[margin=1in]{geometry}

\author{Dan Stelljes}
\date{September 1, 2016}
\title{Previous Paper Analysis}

\begin{document}

\maketitle

\begin{thebibliography}{9}

\bibitem{snyder2012}
  Tim Snyder,
  \emph{Overview and Comparison of Genome Compression Algorithms},
  UMM CSci Senior Seminar Conference, Fall 2012, Morris, MN.

\end{thebibliography}

\end{document}

### 2016-08-29 previous-paper-analysis/summary.tex
Wrote previous paper summary

@@
 
+For this summary, I chose Tim Snyder's \emph{Overview and Comparison of Genome Compression Algorithms} from the Fall 2012 senior seminar conference.
+The paper is a survey of some common methods by which genomic data can be compressed.
+Snyder approaches the topic sensibly: He begins with an introduction to the problem (large amounts of genomic data being produced quickly) and a description of what genomic data actually looks like, discusses some general purpose compression algorithms, and then moves into more specific algorithms for coding single genomes and larger genetic databases.
+
+The structure of the paper is clear throughout.
+Sections are split up appropriately and paragraphs  each deal with a single idea.
+That said, some paragraphs become lengthy and run on, especially when providing examples.
+This excerpt makes up only about a sixth of the paragraph from which it was taken:
+\begin{quote}
+Then, it stores the differences between the two blocks as a string, $x$, of $1$'s and $0$'s where a $1$ at position $i$ means that they are different at position $i$ and a $0$ at position $i$ means that they are the same at position $i$.
+Then arithmetic coding is used to store distance for the differences.
+An example would be if at some index the old block has $A$ and the new block has $G$, the distance from $A$ to $G$ might be $2$.
+\cite{snyder2012}
+\end{quote}
+
+Snyder generally draws from one paper per algorithm.
+For example, the section on Heath's Algorithm is based entirely on the paper in which the algorithm is introduced.
+In addition to papers that describe a particular algorithm, Snyder cites three surveys of compression algorithms, an article promoting cloud computing in genome informatics, and three Wikipedia entries for supporting information.
+
+The only paper that stands out for its age is from 1969 and is used for a supporting comparison (Boulton and Wallace, \emph{The Information Content of a Multistate Distribution}).
+With the exception of a 1994 article also used for supporting information (Grumbach and Tahi, \emph{A New Challenge for Compression Algorithms: Genetic Sequences}), all other sources were published in 2000 or later.
+The papers describing the various compression algorithms are peer-reviewed and found either in conference proceedings or journals.
+Therefore, it seems safe to say that the paper satisfies the expectation that key sources be recent and peer-reviewed.
+
+Snyder incorporates several different methods of explanation throughout the paper.
+All of the algorithms are described in prose.
+Some, as discussed earlier, include examples.
+Snyder also borrows figures from the cited papers to illustrate the database compression algorithms and provides diagrams and pseudocode for the general compression algorithms.
+Overall, the different presentations complement each other well.
+
+The paper flows nicely despite some occasional awkward phrasing.
+Tenses and voices are occasionally confused, and attempts to vary sentence flow are sometimes forced.
+(``I will explain two single genome compression algorithms'' and ``This subsection will explain the Expert Model algorithm'' exist side by side.) Snyder points out obvious flaws in some of the algorithms, but could do a better job of explaining their strengths and weaknesses.
+
+As a whole, the paper was an accessible and (seemingly) complete introduction to the various methods used for genomic compression.
+Snyder presents the information in a clear and easy-to-follow format with appropriate figures, and that aspect is definitely worth emulating.
+
 \begin{thebibliography}{9}

### 2016-09-09 watch.sh
Tried to do annotated bibliography but ended up just dicking around with BibTeX

@@ -1,0 +1,30 @@
+#!/usr/bin/env bash
+
+# adapted from http://superuser.com/a/634313
+# usage: watch.sh annotated-bibliography
+
+# note that the bibliography has to be symlinked to the document directory
+# because BibTeX is a steaming pile of crap:
+DOCUMENT="$1.tex"
+LIBRARY="references.bib"
+
+# move to the document directory because it's easier than trying to get pdflatex
+# to change its output location:
+base=`dirname "$0"`
+cd "$base/$1"
+
+while true; do
+    current=`stat -c %Z "$DOCUMENT" "../$LIBRARY"`
+
+    if [[ "$last" != "$current" ]]; then
+        # we've climbed Everest, run a sub-4 mile, created pizza rolls, and put
+        # people on the moon, but this is apparently still necessary:
+        pdflatex "$DOCUMENT"
+        bibtex `basename -s .tex "$DOCUMENT"`
+        pdflatex "$DOCUMENT"
+        pdflatex "$DOCUMENT"
+
+        last="$current"
+    fi
+    sleep 4
+done

### 2016-09-13 annotated-bibliography/annotated-bibliography.tex
Tried to do annotated bibliography but ended up just dicking around with BibTeX
Brought bibliography in line with UMM style

--- 2016-08-28 previous-paper-analysis/summary.tex
@@
-\documentclass[12pt,letterpaper]{article}
-\usepackage[margin=1in]{geometry}
+\documentclass{sig-alternate}
+\usepackage{url}
 
-\author{Dan Stelljes}
-\date{September 1, 2016}
-\title{Previous Paper Analysis}
+\author{
+\alignauthor{}
+Dan Stelljes\\
+  \affaddr{Division of Science and Mathematics}\\
+  \affaddr{University of Minnesota, Morris}\\
+  \affaddr{Morris, Minnesota, USA 56267}\\
+  \email{stell124@morris.umn.edu}
+}
+\conferenceinfo{UMM CSci Senior Seminar Conference, December 2016}{Morris, MN}
+\title{Composable Concurrency Abstractions}
+\subtitle{Annotated Bibliography}
 
@@
 
-\begin{thebibliography}{9}
+\begin{abstract}
+A wide variety of concurrency models presents challenges in both performance and composability.
+By identifying common concepts and techniques, a high-level language virtual machine can support a variety of abstractions at a low level, and different models can be used in tandem with each other.
+\end{abstract}
 
-\bibitem{snyder2012}
-  Tim Snyder,
-  \emph{Overview and Comparison of Genome Compression Algorithms},
-  UMM CSci Senior Seminar Conference, Fall 2012, Morris, MN.
+\nocite{*}
 
-\end{thebibliography}
+\bibliographystyle{abbrv}
+\bibliography{references}
 

### 2016-09-15 annotated-bibliography/annotated-bibliography.tex
Added new finds to bibliography

@@
 \begin{abstract}
-A wide variety of concurrency models presents challenges in both performance and composability.
-By identifying common concepts and techniques, a high-level language virtual machine can support a variety of abstractions at a low level, and different models can be used in tandem with each other.
+Varied concurrency models and implementations present challenges in both performance and composability.
+By identifying common concepts and techniques, a high-level language virtual machine can support a variety of abstractions at a low level, and different models can be used in tandem.
 \end{abstract}

### 2016-09-18 paper/paper.tex
Started outlining paper

\documentclass{sig-alternate}
\usepackage{biblatex}

\author{
\alignauthor{}
Dan Stelljes\\
  \affaddr{Division of Science and Mathematics}\\
  \affaddr{University of Minnesota, Morris}\\
  \affaddr{Morris, Minnesota, USA 56267}\\
  \email{stell124@morris.umn.edu}
}
\conferenceinfo{UMM CSci Senior Seminar Conference, December 2016}{Morris, MN}
\title{Composable Concurrency Abstractions}

\bibliography{references}

% space is limited:
\AtEveryBibitem{\clearfield{doi}}
\AtEveryBibitem{\clearfield{isbn}}
\AtEveryBibitem{\clearfield{note}}
\AtEveryBibitem{\clearfield{url}}

\begin{document}

\maketitle

\begin{abstract}
Varied concurrency models and implementations present challenges in both performance and composability. By identifying common concepts and techniques, a high-level language virtual machine can support a variety of abstractions at a low level, and different models can be used in tandem.
\end{abstract}

\keywords{concurrency abstraction, concurrency, parallelism, virtual machines}

\section{Introduction}

Concurrency means running multiple tasks in overlapping time periods. It's desirable for a couple of important reasons: complex tasks can be completed faster by splitting them into simple parts that can run concurrently, and high-latency tasks can execute without blocking other tasks.

Concurrent tasks can be controlled in different ways (event loops, software transactional memory, etc.). Someone writing a program might like to mix different control mechanisms to take advantage of particular performance or semantic traits.

\section{Abstraction requirements}

Ziv et al.~\cite{Ziv2015} formalized desired requirements for concurrency abstractions to define a protocol for composition.

\subsection{Serializability}

A history of concurrent operations is serializable if its outcome is the same as the outcome of the operations executed consecutively.

\subsection{Deadlock avoidance}

Progress guarantees avoid deadlock, the state in which two concurrent operations are stuck waiting for the other to finish.

\subsection{Opacity}

A concurrency protocol is opaque if it ensures that each concurrent operation sees a consistent state.

\section{Runtime requirements}

Marr and D'Hondt~\cite{Marr2012} surveyed concurrent programming concepts with the goal of improving support at the virtual machine level. Such support could improve performance and enforce semantic guarantees between languages with a common virtual machine.

\subsection{Managed mutation and execution}

To ensure serializability, any modification of an object's state must be regulated.

\subsection{Ownership}

Mutation and execution must be regulated by some owning entity.

\subsection{Leveled reflection}

Safe language-level reflection and unsafe ``meta-level'' reflection need to be distinguished.

\subsection{Enforcability}

All rules need to be enforceable across all concurrency models.

\section{Conclusions}

\section{Acknowledgments}

\section{References}

\printbibliography[heading=none]

\end{document}

### 2016-09-18 watch.sh

@@ -20,7 +20,7 @@ while true; do
         # we've climbed Everest, run a sub-4 mile, created pizza rolls, and put
         # people on the moon, but this is apparently still necessary:
         pdflatex "$DOCUMENT"
-        bibtex `basename -s .tex "$DOCUMENT"`
+        biber `basename -s .tex "$DOCUMENT"`
         pdflatex "$DOCUMENT"
         pdflatex "$DOCUMENT"
 

### 2016-09-22 paper/paper.tex
Reworked paper structure

@@
 \usepackage{biblatex}
+
+\newcommand{\code}[1]{\texttt{#1}}
 
@@
 \conferenceinfo{UMM CSci Senior Seminar Conference, December 2016}{Morris, MN}
-\title{Composable Concurrency Abstractions}
+\title{Composable Concurrency Models}
 
@@
 \begin{abstract}
-Varied concurrency models and implementations present challenges in both performance and composability.
-By identifying common concepts and techniques, a high-level language virtual machine can support a variety of abstractions at a low level, and different models can be used in tandem.
+
+The need to manage concurrent operations in applications has led to the development of a wide variety of concurrency models.
+Modern programming languages each provide several concurrency models to serve different requirements, and programmers benefit from being able to use them in tandem.
+We discuss challenges surrounding concurrency control and examine situations in which conflicts between models can occur.
+
 \end{abstract}
 
-\keywords{concurrency abstraction, concurrency, parallelism, virtual machines}
+\section*{Key points}
+
+The goal of this research is to discover and explain common properties of concurrency models (the actor model, software transactional memory, etc.) and demonstrate possible conflicts between them, ultimately showing how they can work together.
+The research is scoped to concurrency management---while some of the findings may apply to parallel programming, the source material primarily deals with language-level constructs.
+
+Swalens et al.~\cite{Swalens2014} specifically examine the composability of various concurrency models in Clojure.
+They provide a significant amount of accessible background material, but stop short of actually defining any composable language-level abstractions; the paper serves largely as a foundation for future work.
+Marr (an author on that paper) and D'Hondt~\cite{Marr2012} discuss unifying concurrency abstractions with the goal of increasing virtual machine-level support for various high-level concurrency models.
+They propose an ownership-based meta-object protocol as a solution, and demonstrate the implementation of some common concurrency models.
+Ziv et al.~\cite{Ziv2015} provide formal definitions of requirements for concurrency control (deadlock safety, serializability, etc.) in order to prove composability.
+The paper was written with databases in mind and is (perhaps obviously) quite heavy on theory, but it provides precise definitions for several important concepts.
+
+All of the mentioned sources deal specifically with composability, though approaches are different.
+Swalens et al.~\cite{Swalens2014} work at the language level with a focus on the programmer; Marr and D'Hondt~\cite{Marr2012} work at the virtual machine level with a focus on performant abstractions that can accommodate models from various higher-level languages.
+Taken together, those papers provide a decent foundation for examining different concurrency models.
+Ziv et al.~\cite{Ziv2015} attack the problem from a purely theoretical perspective which, although helpful, doesn't relate particularly well to the more practical approaches.
+
+Swalens et al.~\cite{Swalens2014} demonstrate issues in the context of Clojure's concurrency models, providing code where appropriate.
+Those and similar examples could be quite useful in illustrating composability challenges.
+Marr and D'Hondt~\cite{Marr2012} implement Clojure's agents in Smalltalk and compare that implementation and an implementation of a software transactional memory system (LRSTM) to the proposed meta-object protocol.
+The comparison's benchmark results demonstrate that VM support is necessary to reduce performance of the MOP to an acceptable level, but are not especially relevant outside of that context.
+As far as the general state of the research, it seems that a variety of unifying mechanisms have been proposed but languages generally just work around possible issues (for instance, Clojure doesn't send a message to an agent inside a \code{dosync} block until the transaction is committed).
+
+It seems reasonable to assume that the audience already has an intuitive understanding of different types of concurrent operations (disk reads, file downloads, etc.).
+However, concepts like atomicity and deadlock avoidance may have to be explained.
 
@@
 
-Concurrency means running multiple tasks in overlapping time periods.
-It's desirable for a couple of important reasons: complex tasks can be completed faster by splitting them into simple parts that can run concurrently, and high-latency tasks can execute without blocking other tasks.
+\section{Background}
 
-Concurrent tasks can be controlled in different ways (event loops, software transactional memory, etc.).
-Someone writing a program might like to mix different control mechanisms to take advantage of particular performance or semantic traits.
+\section{Common concurrency models}
 
-\section{Abstraction requirements}
+\subsection{Atomics}
 
-Ziv et al.~\cite{Ziv2015} formalized desired requirements for concurrency abstractions to define a protocol for composition.
+\subsection{Communicating threads}
 
-\subsection{Serializability}
+\subsection{Coroutines}
 
-A history of concurrent operations is serializable if its outcome is the same as the outcome of the operations executed consecutively.
+\subsection{Futures}
 
-\subsection{Deadlock avoidance}
+\subsection{Software transactional memory}
 
-Progress guarantees avoid deadlock, the state in which two concurrent operations are stuck waiting for the other to finish.
+\section{Composability challenges}
 
-\subsection{Opacity}
-
-A concurrency protocol is opaque if it ensures that each concurrent operation sees a consistent state.
-
-\section{Runtime requirements}
-
-Marr and D'Hondt~\cite{Marr2012} surveyed concurrent programming concepts with the goal of improving support at the virtual machine level.
-Such support could improve performance and enforce semantic guarantees between languages with a common virtual machine.
-
-\subsection{Managed mutation and execution}
-
-To ensure serializability, any modification of an object's state must be regulated.
-
-\subsection{Ownership}
-
-Mutation and execution must be regulated by some owning entity.
-
-\subsection{Leveled reflection}
-
-Safe language-level reflection and unsafe ``meta-level'' reflection need to be distinguished.
-
-\subsection{Enforcability}
-
-All rules need to be enforceable across all concurrency models.
+\section{Proposed solutions}
 

### 2016-09-22 paper/paper.tex
Finished outline, sort of

@@
 
-The goal of this research is to discover and explain common properties of concurrency models (the actor model, software transactional memory, etc.) and demonstrate possible conflicts between them, ultimately showing how they can work together.
+The goal of this research is to discover and explain common properties of concurrency models and demonstrate possible conflicts between them, ultimately showing how they can work together.
 The research is scoped to concurrency management---while some of the findings may apply to parallel programming, the source material primarily deals with language-level constructs.
@@
 It seems reasonable to assume that the audience already has an intuitive understanding of different types of concurrent operations (disk reads, file downloads, etc.).
-However, concepts like atomicity and deadlock avoidance may have to be explained.
+However, concurrent programming concepts (atomicity, deadlock, etc.) may have to be explained.
 
@@
 
+Explanation of why concurrency is desirable (and usually necessary in practice). Example of an application that relies heavily on concurrency (web browser, for instance).
+
 \section{Background}
+
+Formal definitions from Ziv et.\ al~\cite{Ziv2015} are likely to be useful (serializability and visibility are treated particularly thoroughly).
+Marr and D'Hondt~\cite{Marr2012}, approaching the topic from a semantic enforcement angle, discuss atomicity in terms of managed mutation and execution.
+Other supplementary resources to fill out basic concurrency concepts would be helpful.
+
+\subsection{Concurrency}
+
+Brief introduction to threads, processes, and cores. Possible discussion of concurrency primitives (coroutines, fibers, green threads).
+
+\subsection{Atomicity}
+
+Explanation of individual effects/mutual exclusion.
+
+\subsection{Serializability}
+
+Explanation of execution order as a requirement for correctness.
+
+\subsection{Visibility}
+
+Explanation of thread opacity and communication.
 
@@
 
+Swalens et al.~\cite{Swalens2014} provide some good explanations of the following, but they may be too closely coupled to Clojure.
+Other sources may be useful for fleshing out definitions or exploring other models.
+
 \subsection{Atomics}
+
+Definition of atomics and atomic operations (compare-and-swap, fetch-and-increment, locking).
 
@@
 
-\subsection{Coroutines}
+Description of concurrency models that use message passing for thread communication rather than shared memory (actors, agents).
 
@@
 
+Explanation of futures (and deferreds and promises) as proxies for some unknown result.
+
 \subsection{Software transactional memory}
 
-\section{Composability challenges}
+Definition of software transactional memory as a model that allows concurrent tasks to write to shared memory.
+
+\section{Composability criteria}
+
+Ziv et al.~\cite{Ziv2015} provide specific formal defininitions for composability correctness (``safety'' is described more specifically in terms of abortable and abort-free serializability and ``liveness'' is referred to as deadlock avoidance).
+Other sources seem to prefer the safety and liveness generalizations.
+
+\subsection{Safety}
+
+Definition of safety as a guarantee of partial correctness (``nothing bad will happen'').
+
+\subsection{Liveness}
+
+Definition of liveness as a guarantee of progress and eventual termination (``something good will eventually happen'').
 
@@
 
+Sources tend to survey a particular space and evaluate solutions within it. This section may require more sources or a narrower scope.
+
+\subsection{Language-specific assistance}
+
+Languages can attempt to prevent common errors. (This maybe isn't as muh a solution as a workaround---Clojure, for instance, will handle some potentially unsafe combinations correctly.)
+
+\subsection{Low-level abstractions}
+
+Lower-level concurrency protocols can serve as the basis of implementation for some models, simplifying composition.
+
 \section{Conclusions}
+
+Eloquent restatement of previous material.
 
@@
 
+Mom, the Academy, etc.
+
 \section{References}
+
+\nocite{*}
 

### 2016-09-29 paper/paper.tex
Panic-wrote section draft

--- 2016-09-18 paper/paper.tex
@@
 \usepackage{biblatex}
+\usepackage{listings}
+
+\newcommand{\code}[1]{\texttt{#1}}
 
@@
 \begin{abstract}
-Varied concurrency models and implementations present challenges in both performance and composability.
-By identifying common concepts and techniques, a high-level language virtual machine can support a variety of abstractions at a low level, and different models can be used in tandem.
+
+The need to manage concurrent operations in applications has led to the development of a wide variety of concurrency models.
+Modern programming languages each provide several concurrency models to serve different requirements, and programmers benefit from being able to use them in tandem.
+We discuss challenges surrounding concurrency control and examine situations in which conflicts between models can occur.
+
 \end{abstract}
-
-\keywords{concurrency abstraction, concurrency, parallelism, virtual machines}
 
@@
 
-Concurrency means running multiple tasks in overlapping time periods.
-It's desirable for a couple of important reasons: complex tasks can be completed faster by splitting them into simple parts that can run concurrently, and high-latency tasks can execute without blocking other tasks.
+Explanation of why concurrency is desirable (and usually necessary in practice). Example of an application that relies heavily on concurrency (web browser, for instance).
 
-Concurrent tasks can be controlled in different ways (event loops, software transactional memory, etc.).
-Someone writing a program might like to mix different control mechanisms to take advantage of particular performance or semantic traits.
+\section{Background}
 
-\section{Abstraction requirements}
+\subsection{Concurrency}
 
-Ziv et al.~\cite{Ziv2015} formalized desired requirements for concurrency abstractions to define a protocol for composition.
+Brief introduction to threads, processes, and cores. Possible discussion of concurrency primitives (coroutines, fibers, green threads).
+
+\subsection{Atomicity}
+
+Explanation of individual effects/mutual exclusion.
 
@@
 
-A history of concurrent operations is serializable if its outcome is the same as the outcome of the operations executed consecutively.
+Explanation of execution order as a requirement for correctness.
 
-\subsection{Deadlock avoidance}
+\subsection{Visibility}
 
-Progress guarantees avoid deadlock, the state in which two concurrent operations are stuck waiting for the other to finish.
+Explanation of thread opacity and communication.
 
-\subsection{Opacity}
+\section{Common concurrency models}
 
-A concurrency protocol is opaque if it ensures that each concurrent operation sees a consistent state.
+\subsection{Atomic variables}
 
-\section{Runtime requirements}
+Atomic variables can only be read and mutated by operations that guarantee atomicity.
+For example, the \emph{compare-and-swap} operation compares the current value of the variable to a given value and only performs a write if the values match:
 
-Marr and D'Hondt~\cite{Marr2012} surveyed concurrent programming concepts with the goal of improving support at the virtual machine level.
-Such support could improve performance and enforce semantic guarantees between languages with a common virtual machine.
+\begin{lstlisting}[language=C]
+  bool swap(int* v, int old, int new) {
+    if (*v != old) {
+      return false;
+    }
 
-\subsection{Managed mutation and execution}
+    *v = new;
+    return true;
+  }
+\end{lstlisting}
 
-To ensure serializability, any modification of an object's state must be regulated.
+\emph{compare-and-swap} ensures that a new value cannot be based on outdated information.
+Suppose that a thread \emph{A} reads a variable \emph{v} and begins computing a new value.
+At the same time, another thread \emph{B} modifies \emph{v}.
+When \emph{A} tries to set the value of \emph{v} via \emph{compare-and-swap}, the write will fail.
+As a result, race conditions are prevented and all writes on \emph{v} are guaranteed to be atomic.
 
-\subsection{Ownership}
+Languages provide varying levels of support for atomic variables.
+Go has no atomic variables, but implements some low-level atomic operations (including \emph{compare-and-swap}).
+Java offers atomic analogues to its primitive wrapper classes (e.g., \code{AtomicInteger} as an atomic \code{Integer}) that ensure thread safety and allow implementations to take advantage of hardware-specific atomic instructions.
+Clojure provides \code{atom} as a reference type; the value of an \code{atom} can be read with \code{deref} and written with \code{swap!}.
 
-Mutation and execution must be regulated by some owning entity.
+\subsection{Software transactional memory}
 
-\subsection{Leveled reflection}
+Software transactional memory (STM) allows multiple concurrent operations to transactionally write to a shared location in memory.
+STM is an example of optimistic concurrency control: Each operation writes to the shared memory with no regard to the activity of other threads.
+After the entire transaction is completed, the transaction manager verifies that other threads have not also made changes to the shared memory.
+If there are conflicting changes, the transaction is aborted and re-executed until it eventually succeeds.~\cite{Shavit1995}
 
-Safe language-level reflection and unsafe ``meta-level'' reflection need to be distinguished.
+Conceptually, STM simplifies concurrency because it allows transactions to be thought of as a single-threaded operation.
+A thread cannot observe changes made to other threads while a transaction is in progress, nor can other threads observe modifications by that thread until the transaction is completed.
+Only when a transaction completes successfully will changes will become visible to other threads.
 
-\subsection{Enforcability}
+Harris and Fraser proposed the use of critical regions to represent transactions:~\cite{Harris2014}
 
-All rules need to be enforceable across all concurrency models.
+\begin{lstlisting}
+  atomic (index > 0) {
+    index--;
+    return buffer[index];
+  }
+\end{lstlisting}
+
+In this (somewhat contrived) example, everything in the \code{atomic} block is executed as a transaction.
+The optional guard condition \code{index > 0} prevents a transaction from occurring if it would fail or is unnecessary.
+No popular languages currently support this type of representation; most often, STM is implemented by a library.
+The Concurrent Ruby library, for example, models transactions with \code{TVar} and \code{Concurrent::atomically}:~\cite{ConcurrentRubyTVar}
+
+\begin{lstlisting}
+  v = TVar.new(50)
+
+  Concurrent::atomically do
+    v.value += 50
+  end
+\end{lstlisting}
+
+\subsection{Communicating threads}
+
+Instead of sharing memory, some concurrency models restrict operations to private memory.
+Threads communicate strictly by message passing, which avoids race conditions.
+Communicating sequential processes (CSP), one of the oldest communicating thread models, describes systems in terms of independent processes that communicate through predefined channels.~\cite{Hoare1978} Other models, most notably the actor model, rely on asynchronous message passing to specific entities.~\cite{Agha1986}
+
+Clojure is useful as a demonstration of communicating threads in that it provides several different models.~\cite{Swalens2014} CSP is implemented by the core.async library.
+\code{chan} opens a new channel and \code{go} creates a new thread.
+Inside a \code{go} block, values can be added to and removed from a channel; subsequent operations will block until another thread performs a complementary operation.
+
+Clojure's agents, by contrast, are asynchronous---there are no blocking receive operations.
+An agent represents the state of an individual entity and only allows changes to occur as the result of an action.
+Actions are applied to the state of an agent and return the new state of the agent.
+
+The actor model is not implemented by Clojure, though it is commonly used in distributed systems.
+(The ``pure'' use of actors assumes that everything is an actor.) Actors can only communicate via asynchronous message passing.
+Upon receiving a message, an actor can choose to send messages to other actors, create new actors, and determine the behavior used for the next received message.
+
+\subsection{Proxies}
+
+Proxies are placeholders for values that are the result of some concurrently executed operation.
+Futures and promises are the two main types of proxy, though the terms are frequently used interchangably (along with ``delayed'' and ``deferred'').
+Specifically, futures are resolved to the result of the completed operation.
+The result is then accessed implicitly; any use of the future will return its value.
+Promises can be created independently and require the result to be accessed explicitly.
+
+JavaScript relies heavily on promises to model concurrency without descending into ``callback hell.'' Rather than deeply nesting callback functions, asynchronous actions can be expressed as promises:
+
+\begin{lstlisting}
+  var read = new Promise((resolve) => {
+    fs.readdir('.', resolve);
+  });
+
+  var count = (files) => {
+    return new Promise((resolve) {
+      resolve(files.length);
+    });
+  };
+
+  var print = console.log;
+
+  read.then(count).then(print);
+\end{lstlisting}
+
+\section{Correctness criteria}
+
+\subsection{Safety}
+
+Definition of safety as a guarantee of partial correctness (``nothing bad will happen'').
+
+\subsection{Liveness}
+
+Definition of liveness as a guarantee of progress and eventual termination (``something good will eventually happen'').
+
+\section{Composability challenges}
+
+Atomics: Operations affecting multiple atomics may cause race conditions. ABA problem.
+
+STM: No irrevocable operations can be performed.
+
+Communicating threads: CSP relies on blocking for communication, and deadlocks are almost inevitable when combined with other models.
+
+Proxies: Reading future/promise values before they're actually available can cause deadlocks. Also, mutually recursive futures.
 
@@
 
+Eloquent restatement of previous material.
+
 \section{Acknowledgments}
 
+Mom, the Academy, etc.
+
 \section{References}
+
+\nocite{*}
 

### 2016-10-10 paper/paper.tex
Fixed references

@@
 \documentclass{sig-alternate}
-\usepackage{biblatex}
-\usepackage{listings}
+\usepackage[backend=bibtex,style=numeric-comp]{biblatex}
 
@@
 
-% space is limited:
-\AtEveryBibitem{\clearfield{doi}}
-\AtEveryBibitem{\clearfield{isbn}}
 \AtEveryBibitem{\clearfield{note}}
-\AtEveryBibitem{\clearfield{url}}
 
@@
 
-\begin{lstlisting}[language=C]
-  bool swap(int* v, int old, int new) {
-    if (*v != old) {
-      return false;
-    }
-
-    *v = new;
-    return true;
-  }
-\end{lstlisting}
-
 \emph{compare-and-swap} ensures that a new value cannot be based on outdated information.
@@
 
-\begin{lstlisting}
-  atomic (index > 0) {
-    index--;
-    return buffer[index];
-  }
-\end{lstlisting}
-
 In this (somewhat contrived) example, everything in the \code{atomic} block is executed as a transaction.
@@
 The Concurrent Ruby library, for example, models transactions with \code{TVar} and \code{Concurrent::atomically}:~\cite{ConcurrentRubyTVar}
-
-\begin{lstlisting}
-  v = TVar.new(50)
-
-  Concurrent::atomically do
-    v.value += 50
-  end
-\end{lstlisting}
 
@@
 
-\begin{lstlisting}
-  var read = new Promise((resolve) => {
-    fs.readdir('.', resolve);
-  });
-
-  var count = (files) => {
-    return new Promise((resolve) {
-      resolve(files.length);
-    });
-  };
-
-  var print = console.log;
-
-  read.then(count).then(print);
-\end{lstlisting}
-
 \section{Correctness criteria}
@@
 
-\section{Acknowledgments}
+\section*{Acknowledgments}
 
@@
 
-\section{References}
-
-\nocite{*}
-
-\printbibliography[heading=none]
+\printbibliography{}
 

### 2016-10-10 watch.sh

@@ -20,7 +20,7 @@ while true; do
         # we've climbed Everest, run a sub-4 mile, created pizza rolls, and put
         # people on the moon, but this is apparently still necessary:
         pdflatex "$DOCUMENT"
-        biber `basename -s .tex "$DOCUMENT"`
+        bibtex `basename -s .tex "$DOCUMENT"`
         pdflatex "$DOCUMENT"
         pdflatex "$DOCUMENT"
 

### 2016-10-10 paper/paper.tex
Cut language examples from models section and wrote introduction

@@
 
+\AtEveryBibitem{\clearfield{doi}}
 \AtEveryBibitem{\clearfield{note}}
@@
 
-The need to manage concurrent operations in applications has led to the development of a wide variety of concurrency models.
-Modern programming languages each provide several concurrency models to serve different requirements, and programmers benefit from being able to use them in tandem.
+The need to manage concurrent operations in applications has led to the development of a variety of concurrency models.
+Modern programming languages generally provide several concurrency models to serve different requirements, and programmers benefit from being able to use them in tandem.
 We discuss challenges surrounding concurrency control and examine situations in which conflicts between models can occur.
+Additionally, we describe attempts to implement concurrency models on top of lower-level concurrency abstractions.
 
@@
 
+\keywords{concurrency, parallelism, concurrency abstraction}
+
 \section{Introduction}
 
-Explanation of why concurrency is desirable (and usually necessary in practice). Example of an application that relies heavily on concurrency (web browser, for instance).
+Most interactive computer programs depend on concurrency, the ability to perform different tasks at the same time.
+A web browser, for instance, might at any point be rendering documents in multiple tabs, transferring files, and handling user interaction.
+On a lower level, the operating system might be running several other applications, juggling background processes, and responding to events.~\cite{Swalens2014} If every long-running task blocked other tasks from proceeding, the system would be effectively unusable.
+
+Concurrency models enable programmers to reason about and describe interactions between concurrent tasks.
+The web browser is a good example of several different models: The user interface layer might rely on an event loop, the rendering process might take place in shared memory, and suggestions from browsing history or a search engine might require parallel collection operations or map-reduce to achieve acceptable performance.~\cite{Marr2012}
+
+Given that an application is likely to make use of more than one concurrency model, programmers would prefer that different models could be combined.
+Such composability is also desirable from an engineering standpoint---a virtual machine for one or more high-level languages should be able to support a variety of models without sacrificing semantics or performance.
+However, different concurrency models are not necessarily composable and unexpected issues may arise when they interact.
+Recent work has attempted to identify unifying concurrency abstractions that would guarantee composability of different models and provide underlying virtual machine support.~\cite{Marr2009, Marr2012, Swalens2014, Ziv2015}
 
@@
 Atomic variables can only be read and mutated by operations that guarantee atomicity.
-For example, the \emph{compare-and-swap} operation compares the current value of the variable to a given value and only performs a write if the values match:
+For example, the \emph{compare-and-swap} operation compares the current value of the variable to a given value and only performs a write if the values match.
 
@@
 As a result, race conditions are prevented and all writes on \emph{v} are guaranteed to be atomic.
-
-Languages provide varying levels of support for atomic variables.
-Go has no atomic variables, but implements some low-level atomic operations (including \emph{compare-and-swap}).
-Java offers atomic analogues to its primitive wrapper classes (e.g., \code{AtomicInteger} as an atomic \code{Integer}) that ensure thread safety and allow implementations to take advantage of hardware-specific atomic instructions.
-Clojure provides \code{atom} as a reference type; the value of an \code{atom} can be read with \code{deref} and written with \code{swap!}.
 
@@
 
-Harris and Fraser proposed the use of critical regions to represent transactions:~\cite{Harris2014}
-
-In this (somewhat contrived) example, everything in the \code{atomic} block is executed as a transaction.
-The optional guard condition \code{index > 0} prevents a transaction from occurring if it would fail or is unnecessary.
-No popular languages currently support this type of representation; most often, STM is implemented by a library.
-The Concurrent Ruby library, for example, models transactions with \code{TVar} and \code{Concurrent::atomically}:~\cite{ConcurrentRubyTVar}
+Harris and Fraser proposed the use of critical regions to represent transactions.~\cite{Harris2014}
 
@@
 Threads communicate strictly by message passing, which avoids race conditions.
-Communicating sequential processes (CSP), one of the oldest communicating thread models, describes systems in terms of independent processes that communicate through predefined channels.~\cite{Hoare1978} Other models, most notably the actor model, rely on asynchronous message passing to specific entities.~\cite{Agha1986}
+Communicating sequential processes (CSP), one of the oldest communicating thread models, describes systems in terms of independent processes that communicate through predefined channels.~\cite{Hoare1978}
 
-Clojure is useful as a demonstration of communicating threads in that it provides several different models.~\cite{Swalens2014} CSP is implemented by the core.async library.
-\code{chan} opens a new channel and \code{go} creates a new thread.
-Inside a \code{go} block, values can be added to and removed from a channel; subsequent operations will block until another thread performs a complementary operation.
-
-Clojure's agents, by contrast, are asynchronous---there are no blocking receive operations.
-An agent represents the state of an individual entity and only allows changes to occur as the result of an action.
-Actions are applied to the state of an agent and return the new state of the agent.
-
-The actor model is not implemented by Clojure, though it is commonly used in distributed systems.
-(The ``pure'' use of actors assumes that everything is an actor.) Actors can only communicate via asynchronous message passing.
+Other models, most notably the actor model, rely on asynchronous message passing to specific entities.~\cite{Agha1986} Actors can only communicate via asynchronous message passing.
 Upon receiving a message, an actor can choose to send messages to other actors, create new actors, and determine the behavior used for the next received message.
@@
 Promises can be created independently and require the result to be accessed explicitly.
-
-JavaScript relies heavily on promises to model concurrency without descending into ``callback hell.'' Rather than deeply nesting callback functions, asynchronous actions can be expressed as promises:
 
@@
 
-\section{Composability challenges}
+\section{Proposed abstractions}
 
-Atomics: Operations affecting multiple atomics may cause race conditions. ABA problem.
+More sources here would be good---Marr and D'Hondt~\cite{Marr2012} provide the only solid abstraction.
 
-STM: No irrevocable operations can be performed.
+\subsection{Asynchronous events}
 
-Communicating threads: CSP relies on blocking for communication, and deadlocks are almost inevitable when combined with other models.
+Ziarek et al.~\cite{Ziarek2011} introduce asynchronous events as a possible abstraction. Their examples are heavily tied to Concurrent ML and might be difficult to integrate.
 
-Proxies: Reading future/promise values before they're actually available can cause deadlocks. Also, mutually recursive futures.
+\subsection{Ownership-based meta-object protocol}
 
-\section{Conclusions}
-
-Eloquent restatement of previous material.
-
-\section*{Acknowledgments}
-
-Mom, the Academy, etc.
+Marr and D'Hondt~\cite{Marr2012} proposed an ownership-based meta-object protocol with the goal of providing VM support. They also demonstrate implementation of actors, STM, agents, and CSP.
 

### 2016-10-13 paper/paper.tex
Started background and descended into quiet panic

@@
 A web browser, for instance, might at any point be rendering documents in multiple tabs, transferring files, and handling user interaction.
-On a lower level, the operating system might be running several other applications, juggling background processes, and responding to events.~\cite{Swalens2014} If every long-running task blocked other tasks from proceeding, the system would be effectively unusable.
+On a lower level, the operating system might be running several other applications, juggling background processes, and responding to events.~\cite{Swalens2014} If every long-running process blocked other processes from proceeding, the system would be effectively unusable.
 
-Concurrency models enable programmers to reason about and describe interactions between concurrent tasks.
-The web browser is a good example of several different models: The user interface layer might rely on an event loop, the rendering process might take place in shared memory, and suggestions from browsing history or a search engine might require parallel collection operations or map-reduce to achieve acceptable performance.~\cite{Marr2012}
+Concurrency models enable programmers to reason about and describe interactions between concurrent processes.
+The web browser is a good example of several different models: The user interface layer might rely on an event loop, the rendering process might operate in shared memory, and suggestions from browsing history or a search engine might require parallel collection operations or map-reduce to achieve acceptable performance.~\cite{Marr2012}
 
 Given that an application is likely to make use of more than one concurrency model, programmers would prefer that different models could be combined.
-Such composability is also desirable from an engineering standpoint---a virtual machine for one or more high-level languages should be able to support a variety of models without sacrificing semantics or performance.
+Composability is also desirable from an engineering standpoint---a virtual machine for one or more high-level languages should be able to support a variety of models without sacrificing semantics or performance.
 However, different concurrency models are not necessarily composable and unexpected issues may arise when they interact.
-Recent work has attempted to identify unifying concurrency abstractions that would guarantee composability of different models and provide underlying virtual machine support.~\cite{Marr2009, Marr2012, Swalens2014, Ziv2015}
+Recent work has attempted to identify unifying concurrency abstractions that would guarantee composability of different models and also provide underlying virtual machine support.~\cite{Marr2009, Marr2012, Swalens2014, Ziv2015}
 
@@
 
+Modern operating systems are expected to run many processes concurrently, and processes themselves are often composed of multiple concurrent threads of execution.
+A processor can only execute one thread at a time, so multitasking is accomplished by rapidly switching between threads.~\cite{Liu1973} Although concurrent threads may appear to be executed simultaneously, truly parallel execution can only take place across multiple processors.
+
+Concurrency models can abstract over implementation details, allowing programmers to reason in terms of asynchronous tasks and independent components instead of low-level thread management.
+In addition to making a program more understandable, higher-level models afford a degree of flexbility: Concurrent tasks could be run in an event loop on one thread, run in multiple threads on the same processor, or run in parallel on multiple processors.
+
 \subsection{Concurrency}
 
-Brief introduction to threads, processes, and cores. Possible discussion of concurrency primitives (coroutines, fibers, green threads).
+For tasks to be executed concurrently, they must be able to be executed out of order or in partial order.
+Lamport, in his foundational work on distributed systems, formalized this by defining a ``happens before'' relation (denoted by ``$\rightarrow$'') on a set of tasks: If $A$ and $B$ are tasks in the same process and $A$ occurs before $B$, or if $A$ is the sending of a message by one process and $B$ is the receipt of the same message by another process, then $A \rightarrow B$.
+Two tasks $A$ and $B$ are said to be concurrent if $A \nrightarrow B$ and $B \nrightarrow A$.~\cite{Lamport1978}
 
-\subsection{Atomicity}
+While the ``happens before'' relation can be used to determine whether tasks can be executed concurrently, it does not imply that the execution will be successful.
+Tasks that depend on shared resources can lead to resource starvation or deadlock, states in which execution cannot continue because a task cannot secure the needed resources.
+A common example is the dining philosophers problem: Five silent philosophers, each with a bowl of spaghetti, sit at a round table.
+A waiter places forks between each pair of adjacent philosophers.
+A philosopher can only eat when he has both left and right forks, and each fork can only be held by one philosopher.~\cite{Hoare1978} Designing an algorithm such that no philosopher will starve is nonobvious; if each philosopher is instructed to pick up the left fork as soon as it becomes available and then pick up the right fork as soon as it becomes available, each philosopher will pick up the fork to his left and wait (forever) for the fork on his right to become available.
+
+\subsection{Linearizability}
 
@@
 For example, the \emph{compare-and-swap} operation compares the current value of the variable to a given value and only performs a write if the values match.
-
 \emph{compare-and-swap} ensures that a new value cannot be based on outdated information.
-Suppose that a thread \emph{A} reads a variable \emph{v} and begins computing a new value.
-At the same time, another thread \emph{B} modifies \emph{v}.
-When \emph{A} tries to set the value of \emph{v} via \emph{compare-and-swap}, the write will fail.
-As a result, race conditions are prevented and all writes on \emph{v} are guaranteed to be atomic.
+Suppose that a thread $A$ reads a variable $v$ and begins computing a new value.
+At the same time, another thread $B$ modifies $v$.
+When $A$ tries to set the value of $v$ via \emph{compare-and-swap}, the write will fail.
+As a result, all writes on $v$ are guaranteed to be atomic.
 

### 2016-10-20 paper/paper.tex
Wrote concurrency background section

--- 2016-10-10 paper/paper.tex
@@
 \usepackage[backend=bibtex,style=numeric-comp]{biblatex}
+\usepackage{tikz}
 
@@
 A web browser, for instance, might at any point be rendering documents in multiple tabs, transferring files, and handling user interaction.
-On a lower level, the operating system might be running several other applications, juggling background processes, and responding to events.~\cite{Swalens2014} If every long-running task blocked other tasks from proceeding, the system would be effectively unusable.
+On a lower level, the operating system might be running several other applications, juggling background processes, and responding to events~\cite{Swalens2014}.
+If every long-running process blocked other processes from proceeding, the system would be effectively unusable.
 
-Concurrency models enable programmers to reason about and describe interactions between concurrent tasks.
-The web browser is a good example of several different models: The user interface layer might rely on an event loop, the rendering process might take place in shared memory, and suggestions from browsing history or a search engine might require parallel collection operations or map-reduce to achieve acceptable performance.~\cite{Marr2012}
+Concurrency models enable programmers to reason about and describe interactions between concurrent processes.
+The web browser is a good example of several different models: The user interface layer might rely on an event loop, the rendering process might operate in shared memory, and suggestions from browsing history or a search engine might require parallel collection operations or map-reduce to achieve acceptable performance~\cite{Marr2012}.
 
 Given that an application is likely to make use of more than one concurrency model, programmers would prefer that different models could be combined.
-Such composability is also desirable from an engineering standpoint---a virtual machine for one or more high-level languages should be able to support a variety of models without sacrificing semantics or performance.
+Composability is also desirable from an engineering standpoint---a virtual machine for one or more high-level languages should be able to support a variety of models without sacrificing semantics or performance.
 However, different concurrency models are not necessarily composable and unexpected issues may arise when they interact.
-Recent work has attempted to identify unifying concurrency abstractions that would guarantee composability of different models and provide underlying virtual machine support.~\cite{Marr2009, Marr2012, Swalens2014, Ziv2015}
+Recent work has attempted to identify unifying concurrency abstractions that would guarantee composability of different models and also provide underlying virtual machine support~\cite{Marr2009, Marr2012, Swalens2014, Ziv2015}.
 
@@
 
+Modern operating systems are expected to run many processes concurrently, and processes themselves are often composed of multiple concurrent threads of execution.
+A processor can only execute one thread at a time, so multitasking is accomplished by rapidly switching between threads~\cite{Liu1973}.
+Although concurrent threads may appear to be executed simultaneously, truly parallel execution can only take place across multiple processors.
+
+Concurrency models can abstract over implementation details, allowing programmers to reason in terms of asynchronous tasks and independent components instead of low-level thread management.
+In addition to making a program more understandable, higher-level models afford a degree of flexibility: Concurrent operations could be executed in an event loop on one thread, executed in multiple threads on the same processor, or executed in parallel.
+
 \subsection{Concurrency}
 
-Brief introduction to threads, processes, and cores. Possible discussion of concurrency primitives (coroutines, fibers, green threads).
+For operations to be executed concurrently, they must be able to be executed out of order or in partial order.
+Lamport, in his foundational work on distributed systems, formalized this by defining a ``happens before'' relation (denoted by ``$\rightarrow$'') on a set of operations: If $A$ and $B$ are operations in the same process and $A$ occurs before $B$, or if $A$ is the sending of a message by one process and $B$ is the receipt of the same message by another process, then $A \rightarrow B$.
+Two operations $A$ and $B$ are said to be concurrent if $A \nrightarrow B$ and $B \nrightarrow A$~\cite{Lamport1978}.
 
-\subsection{Atomicity}
+While the ``happens before'' relation can be used to determine whether operations can be executed concurrently, it does not guarantee the correctness of the results.
+In a non-concurrent (entirely sequential) program, it would be sufficient to show that the program is correct after the execution of each task and therefore correct upon termination.
 
-Explanation of individual effects/mutual exclusion.
+\begin{figure}
+  \centering
+  \begin{tikzpicture}
+    \draw [|-|] (0,0) node[left=3pt] {$ x $} -- (6,0);
+
+    \draw [<-] (0.5,0.25) -- (0.5,0.75) node[above=3pt] {$ 1 $};
+    \draw [->] (1,0.25) -- (1,0.75) node[above=3pt] {$ 1 $};
+
+    \draw [<-] (2,0.25) -- (2,0.75) node[above=3pt] {$ 2 $};
+    \draw [->] (2.75,0.25) -- (2.75,0.75) node[above=3pt] {$ 2 $};
+    \draw [->] (3.25,0.25) -- (3.25,0.75) node[above=3pt] {$ 2 $};
+
+    \draw [<-] (3.75,0.25) -- (3.75,0.75) node[above=3pt] {$ 3 $};
+
+    \draw [<-] (4.5,0.25) -- (4.5,0.75) node[above=3pt] {$ 4 $};
+    \draw [->] (5.5,0.25) -- (5.5,0.75) node[above=3pt] {$ 4 $};
+  \end{tikzpicture}
+  \caption{A single thread writes to and reads from a single variable.}
+\label{figure:single}
+\end{figure}
+
+Figure~\ref{figure:single} illustrates a simple program in which a single thread reads and writes numbers from a variable $x$.
+(Write operations are denoted by an arrow pointing toward the line and read operations by an arrow pointing away.) The program satisfies an intuitive model of how variables should behave---each time $x$ is read, the value is equal to the value of the most recent write.
+
+\begin{figure}
+  \centering
+  \begin{tikzpicture}
+    \draw [|-|] (0,0) node[left=3pt] {$ x $} -- (6,0);
+
+    \draw [<-] (0.5,-0.25) -- (0.5,-0.75) node[below=3pt] {$ 1 $};
+
+    \draw [<-] (0.75,0.25) -- (0.75,0.75) node[above=3pt] {$ 2 $};
+    \draw [->] (1,-0.25) -- (1,-0.75) node[below=3pt] {$ 2 $};
+    \draw [->] (1.5,0.25) -- (1.5,0.75) node[above=3pt] {$ 2 $};
+
+    \draw [<-] (2,-0.25) -- (2,-0.75) node[below=3pt] {$ 3 $};
+    \draw [->] (2.5,0.25) -- (2.5,0.75) node[above=3pt] {$ 3 $};
+    \draw [->] (2.75,-0.25) -- (2.75,-0.75) node[below=3pt] {$ 3 $};
+
+    \draw [<-] (3,0.25) -- (3,0.75) node[above=3pt] {$ 4 $};
+    \draw [->] (3.25,-0.25) -- (3.25,-0.75) node[below=3pt] {$ 4 $};
+
+    \draw [<-] (4,-0.25) -- (4,-0.75) node[below=3pt] {$ 5 $};
+    \draw [->] (4.25,0.25) -- (4.25,0.75) node[above=3pt] {$ 5 $};
+    \draw [->] (4.5,-0.25) -- (4.5,-0.75) node[below=3pt] {$ 5 $};
+
+    \draw [<-] (5,0.25) -- (5,0.75) node[above=3pt] {$ 6 $};
+    \draw [->] (5.25,-0.25) -- (5.25,-0.75) node[below=3pt] {$ 6 $};
+    \draw [->] (5.5,0.25) -- (5.5,0.75) node[above=3pt] {$ 6 $};
+  \end{tikzpicture}
+  \caption{Two threads concurrently write to and read from a single variable.}
+\label{figure:multiple}
+\end{figure}
+
+In a concurrent program, the history of operations (that is, the sequence in which operations are executed) may not be the same for every execution.
+As a result, assumptions about correctness that are true for a sequential program may be violated in a concurrent program.
+This is illustrated in Figure~\ref{figure:multiple}, in which two threads concurrently write from and read to a variable $x$ concurrently.
+Because the operations of the threads are interleaved, a read operation on $x$ may not yield a value that matches the value of that thread's most recent write.
+The value may even be inconsistent between consecutive reads.
+(Errors that arise due to unintentional dependency on execution order are referred to as race conditions.) From the perspective of a single thread, the program appears to be incorrect because it violates the same intuitive model of how variables should behave.
+
+Consistency models guarantee the correctness of a program by defining a set of all allowed histories.
+If an execution of a program follows an allowed history, the execution is consistent.
+If not, the execution is inconsistent.
+If every possible execution of the program follows an allowed history, the program conforms to the consistency model.
+
+\subsection{Linearizability}
+
+Linearizability (also referred to as atomicity or indivisibility) is a guarantee that the completion of a task will appear to be instantaneous.
+In other words, a task will take effect atomically at some point during its execution.
+Any subsequent operation, then, will only see the updated state.
 
@@
 After the entire transaction is completed, the transaction manager verifies that other threads have not also made changes to the shared memory.
-If there are conflicting changes, the transaction is aborted and re-executed until it eventually succeeds.~\cite{Shavit1995}
+If there are conflicting changes, the transaction is aborted and re-executed until it eventually succeeds~\cite{Shavit1995}.
 
@@
 
-Harris and Fraser proposed the use of critical regions to represent transactions.~\cite{Harris2014}
+Harris and Fraser proposed the use of critical regions to represent transactions~\cite{Harris2014}.
 
@@
 Threads communicate strictly by message passing, which avoids race conditions.
-Communicating sequential processes (CSP), one of the oldest communicating thread models, describes systems in terms of independent processes that communicate through predefined channels.~\cite{Hoare1978}
+Communicating sequential processes (CSP), one of the oldest communicating thread models, describes systems in terms of independent processes that communicate through predefined channels~\cite{Hoare1978}.
 
-Other models, most notably the actor model, rely on asynchronous message passing to specific entities.~\cite{Agha1986} Actors can only communicate via asynchronous message passing.
+Other models, most notably the actor model, rely on asynchronous message passing to specific entities~\cite{Agha1986}.
+Actors can only communicate via asynchronous message passing.
 Upon receiving a message, an actor can choose to send messages to other actors, create new actors, and determine the behavior used for the next received message.
@@
 Proxies are placeholders for values that are the result of some concurrently executed operation.
-Futures and promises are the two main types of proxy, though the terms are frequently used interchangably (along with ``delayed'' and ``deferred'').
+Futures and promises are the two main types of proxy, though the terms are frequently used interchangeably (along with ``delayed'' and ``deferred'').
 Specifically, futures are resolved to the result of the completed operation.

### 2016-10-20 paper/paper.tex
Wrote linearizability background section

@@
 While the ``happens before'' relation can be used to determine whether operations can be executed concurrently, it does not guarantee the correctness of the results.
-In a non-concurrent (entirely sequential) program, it would be sufficient to show that the program is correct after the execution of each task and therefore correct upon termination.
+In a non-concurrent (entirely sequential) program, it would be sufficient to show that the program is correct after the execution of each operation and therefore correct upon termination.
+In other words, the program could be proved to be correct by proving that its history (that is, the sequence in which its operations are performed) yields a correct result.
+Consistency models guarantee the correctness of a program by defining a set of all allowed histories~\cite{Ziv2015}.
+If an execution of a program follows an allowed history, the execution is consistent.
+If not, the execution is inconsistent.
+If every possible execution of the program follows an allowed history, the program conforms to the consistency model.
 
-\begin{figure}
+\begin{figure}[h]
   \centering
@@
 
-\begin{figure}
+\begin{figure}[h]
   \centering
@@
 
-In a concurrent program, the history of operations (that is, the sequence in which operations are executed) may not be the same for every execution.
+In a concurrent program, the history of operations may not be the same for every execution.
 As a result, assumptions about correctness that are true for a sequential program may be violated in a concurrent program.
@@
 
-Consistency models guarantee the correctness of a program by defining a set of all allowed histories.
-If an execution of a program follows an allowed history, the execution is consistent.
-If not, the execution is inconsistent.
-If every possible execution of the program follows an allowed history, the program conforms to the consistency model.
-
 \subsection{Linearizability}
 
-Linearizability (also referred to as atomicity or indivisibility) is a guarantee that the completion of a task will appear to be instantaneous.
-In other words, a task will take effect atomically at some point during its execution.
-Any subsequent operation, then, will only see the updated state.
+The examples above assume that all operations complete instantaneously.
+In real systems, though, operations take time.
+Even writing to a location in cache memory, an operation measured in nanoseconds, is not truly instantaneous.
+That an operation may be completed at a time significantly later than the time at which it was invoked introduces uncertainty into a history of operations---the sequence of execution may be influenced by the time it takes for messages to travel.
+
+\begin{figure}[h]
+  \centering
+  \begin{tikzpicture}
+    \draw [|-|] (0,0) node[left=3pt] {$ x $} -- (6,0);
+
+    \draw (0.75,0.25) -- (0.5,0.75) node [above left] {$ 1 $};
+    \draw (0.75,0.25) -- (1,0.75);
+
+    \draw (3.75,-0.25) -- (2,-0.75);
+    \draw (3.75,-0.25) -- (5.5,-0.75) node [below right] {$ 2 $};
+
+    \draw (3.25,0.25) -- (3,0.75) node [above left] {$ 2 $};
+    \draw (3.25,0.25) -- (3.5,0.75);
+  \end{tikzpicture}
+  \caption{Two threads concurrently (and noninstantaneously) write to and read from a single variable.}
+\label{figure:time}
+\end{figure}
+
+Consider a third example (Figure~\ref{figure:time}) in which two threads interact with a variable $x$ noninstantanously.
+(Invocation is denoted by the leftmost point of an event and completion by the rightmost point.
+The point closest to the line indicates the instant at which the operation takes effect.) The lower thread invokes a read while the value of $x$ is set to $1$.
+The upper thread invokes and completes a write between the time that the read is invoked and the value is actually read.
+The read operation then returns $2$ even though $x$ was equal to $1$ at the time it was invoked.
+
+While travel time may introduce ambiguity, there are stil some restrictions on possible sequences of events.
+Specifically, an operation cannot take effect before its invocation, nor can it take effect after its completion.
+Therefore, assuming that all threads share some single global state and that operations on that state take place atomically, each operation will appear to take effect atomically at some point after it is invoked and before it is completed.
+
+This consistency model is referred to as linearizability (also referred to as atomicity or indivisibility), and it guarantees that the completion of an operation will appear to the rest of the system to be instantaneous~\cite{Herlihy1990}.
+In other words, even though linearizable operations are executed concurrently and take time, they appear to happen in a simple linear order.
+
+Linearizability guarantees that the results of an operation will be visible as soon as the operation is complete.
+This can prevent issues such as stale reads (in which a read operation returns a value different from the most recently written value) and non-monotonic reads (in which a later read operation returns an older value than an earlier read operation).
+Linearizability is also a composable guarantee: If all operations in a history are linearizable, then the history itself is linearizable~\cite{Herlihy1990}.
 

### 2016-10-20 paper/paper.tex
Wrote serializability background section

@@
 
-While travel time may introduce ambiguity, there are stil some restrictions on possible sequences of events.
+While travel time may introduce ambiguity, there are still some restrictions on possible sequences of events.
 Specifically, an operation cannot take effect before its invocation, nor can it take effect after its completion.
@@
 
-This consistency model is referred to as linearizability (also referred to as atomicity or indivisibility), and it guarantees that the completion of an operation will appear to the rest of the system to be instantaneous~\cite{Herlihy1990}.
+This consistency model is referred to as linearizability (also referred to as atomicity or indivisibility), and it guarantees that the completion of an operation on a single object will appear to the rest of the system to be instantaneous~\cite{Herlihy1990}.
 In other words, even though linearizable operations are executed concurrently and take time, they appear to happen in a simple linear order.
 
-Linearizability guarantees that the results of an operation will be visible as soon as the operation is complete.
+A useful consequence of linearizability is that the results of an operation must be visible as soon as the operation is complete.
 This can prevent issues such as stale reads (in which a read operation returns a value different from the most recently written value) and non-monotonic reads (in which a later read operation returns an older value than an earlier read operation).
-Linearizability is also a composable guarantee: If all operations in a history are linearizable, then the history itself is linearizable~\cite{Herlihy1990}.
 
@@
 
-Explanation of execution order as a requirement for correctness.
+A history of operations is said to be serializable if it is equivalent to a serial (i.e., non-interleaved) ordering of its operations~\cite{Herlihy1990}.
+Serializability is like linearizability in that it demands some linear order of execution.
+However, serializability describes multiple operations over multiple objects instead of single operations on single objects, and it does not impose any wall-clock time constraints on the history.
+This means that operations may occur in any order as long as some serial history exists.
+
+Serializability alone is a fairly weak consistency model.
+Because it does not place bounds on time or order, events may happen out of sequence.
+However, serializability does guarantee isolation~\cite{Haerder1983}.
+In order for histories to be serializable, they must not overlap and cannot interfere with each other.
+
+Together, linearizability and serializability guarantee strict serializability, in which a history is equivalent to a serial execution order and that order corresponds to the execution order in real time.
+Linearizability, then, can actually be described as strict serializability restricted to single operations on single objects~\cite{Herlihy1990}.
 

### 2016-10-20 paper/paper.tex
Fleshed out atomic variables; repeatedly cursed procrastinative tendencies

@@
 
-\subsection{Visibility}
+\section{Common concurrency models}
 
-Explanation of thread opacity and communication.
-
-\section{Common concurrency models}
+Modern programming languages generally provide several different concurrency models, each with their own terminology and specific implementation details.
+It would be impractical to discuss them all, but the most commonly used models can be adequately described by generalization~\cite{Swalens2014}.
 
@@
 Atomic variables can only be read and mutated by operations that guarantee atomicity.
-For example, the \emph{compare-and-swap} operation compares the current value of the variable to a given value and only performs a write if the values match.
-\emph{compare-and-swap} ensures that a new value cannot be based on outdated information.
+For example, an atomic operation \emph{compare-and-swap} compares the current value of a variable to a given value and only performs a write if the values match~\cite{Swalens2014}.
+\emph{compare-and-swap} ensures that a new value cannot be written based on outdated information.
 Suppose that a thread $A$ reads a variable $v$ and begins computing a new value.
@@
 As a result, all writes on $v$ are guaranteed to be atomic.
+
+Atomic operations like \emph{compare-and-swap} (and, similarly, \emph{fetch-and-add} and \emph{test-and-set}) are often implemented in hardware and guarantee linearizability at the lowest level possible.
+As such, atomic variables often serve as the basis for higher-level concurrency control mechanisms such as semaphores and locks, which in turn support the implementation of concurrent data structures.
+(Linearizability is a composable guarantee~\cite{Herlihy1990}; an operation made up of smaller linearizable operations is itself linearizable.)
+
+A counting semaphore is an atomic variable that serves as a record of how many units of a shared resource are available.
+By taking advantage of atomic operations, the semaphore can be safely updated and will reliably determine whether a shared resource can be used.
+A single thread can then test if a resource is available and acquire it before proceeding, thereby preventing race conditions~\cite{Swalens2014}.
+A lock (also referred to as a mutex) guarantees mutual exclusion, which requires that two threads cannot operate on a shared object at the same time.
+Locks are commonly implemented by a binary semaphore (that is, a semaphore that simply indicates whether a single resource is available).
+
+Higher-level languages (such as Clojure~\cite{Swalens2014}) often provide atomic references that can be read by an atomic dereferencing operation and written by an atomic swap operation.
+Such operations may provide convenience features such as automatic retries.
+In practice, atomics are used to share independent objects that do not demand coordinated updates~\cite{Swalens2014}.
 
@@
 
+\section{Conclusion}
+
+In more words: There's a lot of work on the correctness of concurrent operations.
+With some care, concurrency models can be composed.
+There's some work on deeper abstractions, but it's definitely ongoing and more is likely to follow soon.
+
+\section*{Acknowledgments}
+
+Thanks to Elena Machkasova and K.~K. Lamberty for their patience and feedback.
+
 \printbibliography{}

### 2016-10-20 paper/paper.tex
Fleshed out STM; pondered slides

@@
 
-Software transactional memory (STM) allows multiple concurrent operations to transactionally write to a shared location in memory.
-STM is an example of optimistic concurrency control: Each operation writes to the shared memory with no regard to the activity of other threads.
+Software transactional memory (STM) presents an alternative to semaphores and locks by allowing multiple concurrent operations to write to a shared location in memory.
+STM is an example of optimistic concurrency control: Each thread executes a series of read and write operations (called a transaction) on the shared memory without considering the activity of other threads.
+On its face, this approach seems disaster-prone---multiple threads writing to the same memory location could easily lead to corrupted data.
+STM solves this problem by recording all operations in a log.
 After the entire transaction is completed, the transaction manager verifies that other threads have not also made changes to the shared memory.
-If there are conflicting changes, the transaction is aborted and re-executed until it eventually succeeds~\cite{Shavit1995}.
+If there are conflicting changes, the transaction is aborted and retried until it eventually succeeds~\cite{Shavit1995}.
 
-Conceptually, STM simplifies concurrency because it allows transactions to be thought of as a single-threaded operation.
+Conceptually, STM simplifies concurrency because it allows a transaction to be thought of as a single strictly serializable operation.
 A thread cannot observe changes made to other threads while a transaction is in progress, nor can other threads observe modifications by that thread until the transaction is completed.
@@
 
-Harris and Fraser proposed the use of critical regions to represent transactions~\cite{Harris2014}.
+STM also offers significant advantages over lock-based programming in terms of composability.
+Harris et al.\ describe a STM that allows any combination of atomic operations to be composed into a larger atomic operation~\cite{Harris2005}.
+Generally speaking, executing two operations within a transaction yields a larger atomic operation.
+
+However, the requirement that transactions must be able to be aborted and retried places constraints on the set of allowed operations.
+Specifically, a transaction cannot execute any operation that cannot be undone.
+STM can also result in a decrease in performance caused by the overhead of maintaining the log and aborting and retrying transactions.
 

### 2016-10-20 paper/paper.tex
Vomited out some stuff about communicating threads

@@
 Only when a transaction completes successfully will changes will become visible to other threads.
+Practically, STM is useful in any situation in which a shared object needs to be accessed and modified by multiple threads~\cite{Swalens2014}.
 
@@
 
-Instead of sharing memory, some concurrency models restrict operations to private memory.
+Other concurrently models avoid the use of shared memory (and the problems that arise from doing so) by restricting threads to private memory.
 Threads communicate strictly by message passing, which avoids race conditions.
+A variety of different communicating thread models exist and have long been a solution to describing concurrent operations.
+Practical use cases for communicating threads include communication with external systems or running an event loop~\cite{Swalens2014}.
+
 Communicating sequential processes (CSP), one of the oldest communicating thread models, describes systems in terms of independent processes that communicate through predefined channels~\cite{Hoare1978}.
+CSP is notable in that messages are passed synchronously; that is, a sending process will block until a complementary operation is executed on the receiving process~\cite{Swalens2014}.
 
@@
 Upon receiving a message, an actor can choose to send messages to other actors, create new actors, and determine the behavior used for the next received message.
+
+Shared state can also be represented by communicating threads.
+The agent model, for instance, works similarly to atomic references.
+A state wrapped by an agent can be modified by a message that sends an updated state.
+The value can be accessed by a dereferencing operation that reads the current value from the agent~\cite{Swalens2014}.
 
@@
 \section{Correctness criteria}
+
+This section may be unnecessary. There's already a lot of overlapping material in the background section that could be made more comprehensive.
 

### 2016-10-24 slides/slides.tex
Started slide deck (and moved to XeTex)

\documentclass{beamer}

% this actually isn't totally hideous:
\usetheme{metropolis}
\setsansfont{Noto Sans}

\title{Composable Concurrency Models}
\date{November 19, 2016}
\author{Dan Stelljes}
\institute{University of Minnesota Morris}

\begin{document}
  \maketitle

  \begin{frame}{Concurrency Models}
    Example thing goes here.
  \end{frame}
\end{document}

### 2016-10-24 watch.sh

@@ -19,10 +19,10 @@ while true; do
     if [[ "$last" != "$current" ]]; then
         # we've climbed Everest, run a sub-4 mile, created pizza rolls, and put
         # people on the moon, but this is apparently still necessary:
-        pdflatex "$DOCUMENT"
+        xelatex "$DOCUMENT"
         bibtex `basename -s .tex "$DOCUMENT"`
-        pdflatex "$DOCUMENT"
-        pdflatex "$DOCUMENT"
+        xelatex "$DOCUMENT"
+        xelatex "$DOCUMENT"
 
         last="$current"
     fi

### 2016-10-27 paper/paper.tex
Started OBMOP section

@@
 Given that an application is likely to make use of more than one concurrency model, programmers would prefer that different models could be combined.
-Composability is also desirable from an engineering standpoint---a virtual machine for one or more high-level languages should be able to support a variety of models without sacrificing semantics or performance.
+Composability is also desirable from an engineering standpoint---a virtual machine for one or more high-level languages should be able to support a variety of models without resorting to rough low-level approximations.
 However, different concurrency models are not necessarily composable and unexpected issues may arise when they interact.
@@
 
-Marr and D'Hondt~\cite{Marr2012} proposed an ownership-based meta-object protocol with the goal of providing VM support. They also demonstrate implementation of actors, STM, agents, and CSP.
+Marr and D'Hondt~\cite{Marr2012} surveyed a variety of concurrent and parallel programming concepts to identify a unifying concurrency abstraction that could be implemented at the virtual machine (VM) level.\footnote{A 2009 survey by Marr, Haupt, and D'Hondt found that virtual machines at the time did not provide intrinsic support for more than one specific model~\cite{Marr2009}.} A unifying abstraction would enable each concurrency model in a high-level language to be represented in machine code without sacrificing performance or semantics.
+To identify concepts relevant for a multi-language VM, Marr and D'Hondt selected four questions~\cite{Marr2012}:
+
+\begin{enumerate}
+  \item Can the concept be reasonably implemented as a library? (A library implementation may suffer from loss of performance or semantics to an extent that the concept would warrant VM inclusion.)
+
+  \item Does the concept require runtime support to guarantee its semantics? (Semantic guarantees may be enforced by a compiler but not the VM.)
+
+  \item Would runtime support significantly benefit performance? (Deeper knowledge of the concept may enable the VM to better optimize.)
+
+  \item Is the concept already supported by a common VM like the Java Virtual Machine (JVM) or the Common Language Runtime (CLR)?
+\end{enumerate}
+
+From those questions, 26 concepts were identified that would benefit from better low-level support.
 

### 2016-10-27 paper/paper.tex
Added additional figures and incorporated some of Elena's background suggestions

@@
 \usepackage[backend=bibtex,style=numeric-comp]{biblatex}
-\usepackage{tikz}
+\usepackage{graphicx}
+
+\graphicspath{{../figures/}}
 
@@
 Concurrency models enable programmers to reason about and describe interactions between concurrent processes.
-The web browser is a good example of several different models: The user interface layer might rely on an event loop, the rendering process might operate in shared memory, and suggestions from browsing history or a search engine might require parallel collection operations or map-reduce to achieve acceptable performance~\cite{Marr2012}.
+The web browser is a good example of several different models: The user interface layer might rely on an event loop, the rendering process might operate in shared memory, and suggestions from browsing history or a search engine might require parallel collection operations to achieve acceptable performance~\cite{Marr2012}.
 
-Given that an application is likely to make use of more than one concurrency model, programmers would prefer that different models could be combined.
-Composability is also desirable from an engineering standpoint---a virtual machine for one or more high-level languages should be able to support a variety of models without resorting to rough low-level approximations.
+Given that an application is likely to make use of more than one concurrency model, programmers would prefer that different types of models could be arbitrarily combined.
+From an engineering standpoint, composable models are also desirable---a virtual machine for one or more high-level languages should be able to support a variety of models without resorting to rough low-level approximations.
 However, different concurrency models are not necessarily composable and unexpected issues may arise when they interact.
@@
 
-Concurrency models can abstract over implementation details, allowing programmers to reason in terms of asynchronous tasks and independent components instead of low-level thread management.
-In addition to making a program more understandable, higher-level models afford a degree of flexibility: Concurrent operations could be executed in an event loop on one thread, executed in multiple threads on the same processor, or executed in parallel.
+Concurrency models can abstract over implementation details, allowing programmers to reason in terms of asynchron\-ous tasks and independent components instead of low-level thread management.
+In addition to making a program more understandable, higher-level models afford a degree of flexibility: Regardless of whether a set of concurrent operations is executed in sequence on one thread, in multiple threads on the same processor, or in parallel on multiple processors, the result will be the same.
 
@@
 For operations to be executed concurrently, they must be able to be executed out of order or in partial order.
-Lamport, in his foundational work on distributed systems, formalized this by defining a ``happens before'' relation (denoted by ``$\rightarrow$'') on a set of operations: If $A$ and $B$ are operations in the same process and $A$ occurs before $B$, or if $A$ is the sending of a message by one process and $B$ is the receipt of the same message by another process, then $A \rightarrow B$.
-Two operations $A$ and $B$ are said to be concurrent if $A \nrightarrow B$ and $B \nrightarrow A$~\cite{Lamport1978}.
+Lamport, in his foundational work on distributed systems~\cite{Lamport1978}, formalized this by defining a ``happens before'' relation (denoted by ``$\rightarrow$'' and its negation by ``$\nrightarrow$'') on a set of operations: If $A$ and $B$ are operations in the same process and $A$ occurs before $B$, or if $A$ is the sending of a message by one process and $B$ is the receipt of the same message by another process, then $A \rightarrow B$.
+Two operations $A$ and $B$ are said to be concurrent if $A \nrightarrow B$ and $B \nrightarrow A$.
 
@@
 In other words, the program could be proved to be correct by proving that its history (that is, the sequence in which its operations are performed) yields a correct result.
-Consistency models guarantee the correctness of a program by defining a set of all allowed histories~\cite{Ziv2015}.
+Consistency models guarantee the correctness of a program by defining a set of all allowed histories, or histories that lead to a correct result~\cite{Ziv2015}.
 If an execution of a program follows an allowed history, the execution is consistent.
@@
   \centering
-  \begin{tikzpicture}
-    \draw [|-|] (0,0) node[left=3pt] {$ x $} -- (6,0);
-
-    \draw [<-] (0.5,0.25) -- (0.5,0.75) node[above=3pt] {$ 1 $};
-    \draw [->] (1,0.25) -- (1,0.75) node[above=3pt] {$ 1 $};
-
-    \draw [<-] (2,0.25) -- (2,0.75) node[above=3pt] {$ 2 $};
-    \draw [->] (2.75,0.25) -- (2.75,0.75) node[above=3pt] {$ 2 $};
-    \draw [->] (3.25,0.25) -- (3.25,0.75) node[above=3pt] {$ 2 $};
-
-    \draw [<-] (3.75,0.25) -- (3.75,0.75) node[above=3pt] {$ 3 $};
-
-    \draw [<-] (4.5,0.25) -- (4.5,0.75) node[above=3pt] {$ 4 $};
-    \draw [->] (5.5,0.25) -- (5.5,0.75) node[above=3pt] {$ 4 $};
-  \end{tikzpicture}
-  \caption{A single thread writes to and reads from a single variable.}
+  \includegraphics{single-thread-read-write}
+  \caption{A single thread writes to and reads from a variable.}
 \label{figure:single}
@@
 Figure~\ref{figure:single} illustrates a simple program in which a single thread reads and writes numbers from a variable $x$.
-(Write operations are denoted by an arrow pointing toward the line and read operations by an arrow pointing away.) The program satisfies an intuitive model of how variables should behave---each time $x$ is read, the value is equal to the value of the most recent write.
+The value of $x$ over time is represented by a segmented bar.
+Write operations are denoted by an arrow pointing toward the bar and read operations by an arrow pointing away.
+The program satisfies an intuitive model of how variables should behave---each time $x$ is read, the value is equal to the value of the most recent write.
 
@@
   \centering
-  \begin{tikzpicture}
-    \draw [|-|] (0,0) node[left=3pt] {$ x $} -- (6,0);
-
-    \draw [<-] (0.5,-0.25) -- (0.5,-0.75) node[below=3pt] {$ 1 $};
-
-    \draw [<-] (0.75,0.25) -- (0.75,0.75) node[above=3pt] {$ 2 $};
-    \draw [->] (1,-0.25) -- (1,-0.75) node[below=3pt] {$ 2 $};
-    \draw [->] (1.5,0.25) -- (1.5,0.75) node[above=3pt] {$ 2 $};
-
-    \draw [<-] (2,-0.25) -- (2,-0.75) node[below=3pt] {$ 3 $};
-    \draw [->] (2.5,0.25) -- (2.5,0.75) node[above=3pt] {$ 3 $};
-    \draw [->] (2.75,-0.25) -- (2.75,-0.75) node[below=3pt] {$ 3 $};
-
-    \draw [<-] (3,0.25) -- (3,0.75) node[above=3pt] {$ 4 $};
-    \draw [->] (3.25,-0.25) -- (3.25,-0.75) node[below=3pt] {$ 4 $};
-
-    \draw [<-] (4,-0.25) -- (4,-0.75) node[below=3pt] {$ 5 $};
-    \draw [->] (4.25,0.25) -- (4.25,0.75) node[above=3pt] {$ 5 $};
-    \draw [->] (4.5,-0.25) -- (4.5,-0.75) node[below=3pt] {$ 5 $};
-
-    \draw [<-] (5,0.25) -- (5,0.75) node[above=3pt] {$ 6 $};
-    \draw [->] (5.25,-0.25) -- (5.25,-0.75) node[below=3pt] {$ 6 $};
-    \draw [->] (5.5,0.25) -- (5.5,0.75) node[above=3pt] {$ 6 $};
-  \end{tikzpicture}
-  \caption{Two threads concurrently write to and read from a single variable.}
+  \includegraphics{multiple-threads-read-write}
+  \caption{Two threads concurrently write to and read from a variable.}
 \label{figure:multiple}
@@
 The value may even be inconsistent between consecutive reads.
-(Errors that arise due to unintentional dependency on execution order are referred to as race conditions.) From the perspective of a single thread, the program appears to be incorrect because it violates the same intuitive model of how variables should behave.
+If a single thread were to incorrectly assume that it had exclusive control of $x$, this behavior would appear incorrect and could lead to race conditions, errors that arise due to unintentional dependency on execution order are referred to as race conditions.
 
@@
   \centering
-  \begin{tikzpicture}
-    \draw [|-|] (0,0) node[left=3pt] {$ x $} -- (6,0);
-
-    \draw (0.75,0.25) -- (0.5,0.75) node [above left] {$ 1 $};
-    \draw (0.75,0.25) -- (1,0.75);
-
-    \draw (3.75,-0.25) -- (2,-0.75);
-    \draw (3.75,-0.25) -- (5.5,-0.75) node [below right] {$ 2 $};
-
-    \draw (3.25,0.25) -- (3,0.75) node [above left] {$ 2 $};
-    \draw (3.25,0.25) -- (3.5,0.75);
-  \end{tikzpicture}
-  \caption{Two threads concurrently (and noninstantaneously) write to and read from a single variable.}
+  \includegraphics{noninstantaneity}
+  \caption{Two threads concurrently (and noninstantaneously) write to and read from a variable.}
 \label{figure:time}
@@
 Consider a third example (Figure~\ref{figure:time}) in which two threads interact with a variable $x$ noninstantanously.
-(Invocation is denoted by the leftmost point of an event and completion by the rightmost point.
-The point closest to the line indicates the instant at which the operation takes effect.) The lower thread invokes a read while the value of $x$ is set to $1$.
+Invocation is denoted by the leftmost point of an event and completion by the rightmost point.
+The dotted line indicates the instant at which the operation takes effect.
+The lower thread invokes a read while the value of $x$ is set to $1$.
 The upper thread invokes and completes a write between the time that the read is invoked and the value is actually read.
@@
 
+\begin{figure}[h]
+  \centering
+  \includegraphics{linearizability}
+  \caption{A single thread performs an atomic write operation on a variable.}
+\label{figure:linearizability}
+\end{figure}
+
 This consistency model is referred to as linearizability (also referred to as atomicity or indivisibility), and it guarantees that the completion of an operation on a single object will appear to the rest of the system to be instantaneous~\cite{Herlihy1990}.
 In other words, even though linearizable operations are executed concurrently and take time, they appear to happen in a simple linear order.
+Figure~\ref{figure:linearizability} demonstrates an atomic write on a variable $x$: Although the operation takes time, the entire operation (denoted by the shaded area) can be collapsed into one apparently instantaneous event.
 
@@
 This can prevent issues such as stale reads (in which a read operation returns a value different from the most recently written value) and non-monotonic reads (in which a later read operation returns an older value than an earlier read operation).
+Linearizability is also a composable guarantee~\cite{Herlihy1990}---an operation made up of smaller linearizable operations is itself linearizable.
 
@@
 This means that operations may occur in any order as long as some serial history exists.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics{serializability}
+  \caption{Two threads perform a set of serializable read and write operations on a variable.}
+\label{figure:serializability}
+\end{figure}
+
+Figure~\ref{figure:serializability} illustrates a set of serializable read and write operations.
+Even though the durations of each event occur out of order and even cross, the events themselves are equivalent to a serial history: $x \leftarrow 1$, $x = 1$, $x \leftarrow 2$, $x = 2$.
 
@@
 As such, atomic variables often serve as the basis for higher-level concurrency control mechanisms such as semaphores and locks, which in turn support the implementation of concurrent data structures.
-(Linearizability is a composable guarantee~\cite{Herlihy1990}; an operation made up of smaller linearizable operations is itself linearizable.)
 

### 2016-10-27 paper/paper.tex
Fixed some nitpicky things and prepped for serious work abstractions section

@@
 The value may even be inconsistent between consecutive reads.
-If a single thread were to incorrectly assume that it had exclusive control of $x$, this behavior would appear incorrect and could lead to race conditions, errors that arise due to unintentional dependency on execution order are referred to as race conditions.
+If a single thread were to incorrectly assume that it had exclusive control of $x$, this behavior would appear incorrect and could lead to race conditions, errors that arise due to unintentional dependency on execution order.
 
@@
 Figure~\ref{figure:serializability} illustrates a set of serializable read and write operations.
-Even though the durations of each event occur out of order and even cross, the events themselves are equivalent to a serial history: $x \leftarrow 1$, $x = 1$, $x \leftarrow 2$, $x = 2$.
+Even though the durations of each event occur out of order and even cross, the events themselves can be arranged serially: $x \leftarrow 1$, $x = 1$, $x \leftarrow 2$, $x = 2$.
 
 Serializability alone is a fairly weak consistency model.
-Because it does not place bounds on time or order, events may happen out of sequence.
-However, serializability does guarantee isolation~\cite{Haerder1983}.
-In order for histories to be serializable, they must not overlap and cannot interfere with each other.
+Because it does not place bounds on time or order, events may happen out of sequence, and, unlike linearizability, serializability is not a composable guarantee.
+Serializability is useful, though, as a guarantee of isolation~\cite{Haerder1983}: While a serializable set of operations is being executed, it appears to be the only set of operations being executed.
+For histories to be serializable, then, they must not overlap and cannot interfere with each other.
 
-Together, linearizability and serializability guarantee strict serializability, in which a history is equivalent to a serial execution order and that order corresponds to the execution order in real time.
-Linearizability, then, can actually be described as strict serializability restricted to single operations on single objects~\cite{Herlihy1990}.
+Together, linearizability and serializability guarantee strict serializability, in which a history is equivalent to a serial execution order and that serial order corresponds to the execution order in real time.
+Given that, linearizability can actually be described as strict serializability restricted to single operations on single objects~\cite{Herlihy1990}.
 
@@
 
-\section{Correctness criteria}
-
-This section may be unnecessary. There's already a lot of overlapping material in the background section that could be made more comprehensive.
-
-\subsection{Safety}
-
-Definition of safety as a guarantee of partial correctness (``nothing bad will happen'').
-
-\subsection{Liveness}
-
-Definition of liveness as a guarantee of progress and eventual termination (``something good will eventually happen'').
-
 \section{Proposed abstractions}
-
-More sources here would be good---Marr and D'Hondt~\cite{Marr2012} provide the only solid abstraction.
 

### 2016-10-27 paper/paper.tex
Added some stuff about proxies and cleaned up models section

@@
 \documentclass{sig-alternate}
+\usepackage[USenglish]{babel}
 \usepackage[backend=bibtex,style=numeric-comp]{biblatex}
@@
 Given that an application is likely to make use of more than one concurrency model, programmers would prefer that different types of models could be arbitrarily combined.
-From an engineering standpoint, composable models are also desirable---a virtual machine for one or more high-level languages should be able to support a variety of models without resorting to rough low-level approximations.
+From an engineering standpoint, composable models are also desir\-able---a virtual machine for one or more high-level languages should be able to support a variety of models without resorting to rough low-level approximations.
 However, different concurrency models are not necessarily composable and unexpected issues may arise when they interact.
@@
 Figure~\ref{figure:serializability} illustrates a set of serializable read and write operations.
-Even though the durations of each event occur out of order and even cross, the events themselves can be arranged serially: $x \leftarrow 1$, $x = 1$, $x \leftarrow 2$, $x = 2$.
+Even though the exeuctions of each operation occur out of order and even intersect, the events themselves can be arranged serially: $x \leftarrow 1$, $x = 1$, $x \leftarrow 2$, $x = 2$.
 
@@
 
-Atomic variables can only be read and mutated by operations that guarantee atomicity.
-For example, an atomic operation \emph{compare-and-swap} compares the current value of a variable to a given value and only performs a write if the values match~\cite{Swalens2014}.
+Atomic variables are used to share independent objects that do not demand coordinated updates but need to be shared by multiple threads~\cite{Swalens2014}.
+Atomic variables can only be read and mutated by operations that are guaranteed to be linearizable.
+For example, the atomic operation \emph{compare-and-swap} compares the current value of a variable to a given value and only performs a write if the values match~\cite{Swalens2014}.
 \emph{compare-and-swap} ensures that a new value cannot be written based on outdated information.
-Suppose that a thread $A$ reads a variable $v$ and begins computing a new value.
-At the same time, another thread $B$ modifies $v$.
-When $A$ tries to set the value of $v$ via \emph{compare-and-swap}, the write will fail.
-As a result, all writes on $v$ are guaranteed to be atomic.
+Suppose that a thread $A$ reads a variable $x$ and begins computing a new value.
+At the same time, another thread $B$ modifies $x$.
+When $A$ tries to set the value of $x$ via \emph{compare-and-swap}, the write will fail.
+As a result, all writes on $x$ are guaranteed to be linearizable.
 
-Atomic operations like \emph{compare-and-swap} (and, similarly, \emph{fetch-and-add} and \emph{test-and-set}) are often implemented in hardware and guarantee linearizability at the lowest level possible.
-As such, atomic variables often serve as the basis for higher-level concurrency control mechanisms such as semaphores and locks, which in turn support the implementation of concurrent data structures.
+Atomic operations are often implemented in hardware and guarantee linearizability at the lowest level possible.
+\emph{fetch-and-add}, another atomic operation, is a common processor instruction that executes multiple hardware-level operations: The value old value is copied from a location in memory into a temporary register, addition is performed on the value in the temorary register, and the new value is stored at the original location.
 
+Atomic variables often serve as the basis for higher-level concurrency control mechanisms such as semaphores and locks, which in turn support the implementation of concurrent data structures.
 A counting semaphore is an atomic variable that serves as a record of how many units of a shared resource are available.
@@
 
-Higher-level languages (such as Clojure~\cite{Swalens2014}) often provide atomic references that can be read by an atomic dereferencing operation and written by an atomic swap operation.
+Many higher-level languages (such as Clojure~\cite{Swalens2014}) provide atomic references that can be read by an atomic dereferencing operation and written by an atomic swap operation.
 Such operations may provide convenience features such as automatic retries.
-In practice, atomics are used to share independent objects that do not demand coordinated updates~\cite{Swalens2014}.
 
@@
 STM also offers significant advantages over lock-based programming in terms of composability.
-Harris et al.\ describe a STM that allows any combination of atomic operations to be composed into a larger atomic operation~\cite{Harris2005}.
-Generally speaking, executing two operations within a transaction yields a larger atomic operation.
+Given that executing two operations within a transaction yields a larger atomic operation, Harris et al.\ describe a STM that allows any combination of atomic operations to be composed into a larger atomic operation~\cite{Harris2005}.
 
 However, the requirement that transactions must be able to be aborted and retried places constraints on the set of allowed operations.
-Specifically, a transaction cannot execute any operation that cannot be undone.
+Specifically, a transaction cannot execute any operation that cannot be undone or retried, such as writing to disk or performing a network request.
 STM can also result in a decrease in performance caused by the overhead of maintaining the log and aborting and retrying transactions.
@@
 
-Proxies are placeholders for values that are the result of some concurrently executed operation.
-Futures and promises are the two main types of proxy, though the terms are frequently used interchangeably (along with ``delayed'' and ``deferred'').
-Specifically, futures are resolved to the result of the completed operation.
+Proxies are a general term used to describe placeholders for values that are the result of some concurrently executed operation.
+Unlike communicating thread models, which rely on message passing between dedicated threads, a proxy executes some function in a new thread and delivers the result upon completion~\cite{Swalens2014}.
+
+Futures and promises are two popular proxy models, though the terms are frequently used interchangeably (along with ``delayed,'' ``deferred,'' and ``eventual'').
+Generally, futures are resolved to the result of the completed operation.
 The result is then accessed implicitly; any use of the future will return its value.
-Promises can be created independently and require the result to be accessed explicitly.
+Promises, on the other hand, are created as independent objects and require the result to be accessed explicitly.
+
+In practice, proxies are used to execute long-running operations such as rendering or network requests~\cite{Swalens2014} without blocking other operations.
+Promise objects, which can be easily passed and composed, generally offer more flexibility than constructs such as asynchronous callbacks~\cite{Kambona2013}.
 

### 2016-10-27 paper/paper.tex
Finished OBMOP section

@@
 Atomic variables can only be read and mutated by operations that are guaranteed to be linearizable.
-For example, the atomic operation \emph{compare-and-swap} compares the current value of a variable to a given value and only performs a write if the values match~\cite{Swalens2014}.
-\emph{compare-and-swap} ensures that a new value cannot be written based on outdated information.
+For example, the atomic operation \code{compare-and-swap} compares the current value of a variable to a given value and only performs a write if the values match~\cite{Swalens2014}.
+\code{compare-and-swap} ensures that a new value cannot be written based on outdated information.
 Suppose that a thread $A$ reads a variable $x$ and begins computing a new value.
 At the same time, another thread $B$ modifies $x$.
-When $A$ tries to set the value of $x$ via \emph{compare-and-swap}, the write will fail.
+When $A$ tries to set the value of $x$ via \code{compare-and-swap}, the write will fail.
 As a result, all writes on $x$ are guaranteed to be linearizable.
@@
 Atomic operations are often implemented in hardware and guarantee linearizability at the lowest level possible.
-\emph{fetch-and-add}, another atomic operation, is a common processor instruction that executes multiple hardware-level operations: The value old value is copied from a location in memory into a temporary register, addition is performed on the value in the temorary register, and the new value is stored at the original location.
+\code{fetch-and-add}, another atomic operation, is a common processor instruction that executes multiple hardware-level operations: The value old value is copied from a location in memory into a temporary register, addition is performed on the value in the temorary register, and the new value is stored at the original location.
 
@@
 
-Futures and promises are two popular proxy models, though the terms are frequently used interchangeably (along with ``delayed,'' ``deferred,'' and ``eventual'').
+Futures and promises are two of the most common proxy models, though the terms are frequently used interchangeably (along with ``delayed,'' ``deferred,'' and ``eventual'').
 Generally, futures are resolved to the result of the completed operation.
@@
 
-Marr and D'Hondt~\cite{Marr2012} surveyed a variety of concurrent and parallel programming concepts to identify a unifying concurrency abstraction that could be implemented at the virtual machine (VM) level.\footnote{A 2009 survey by Marr, Haupt, and D'Hondt found that virtual machines at the time did not provide intrinsic support for more than one specific model~\cite{Marr2009}.} A unifying abstraction would enable each concurrency model in a high-level language to be represented in machine code without sacrificing performance or semantics.
-To identify concepts relevant for a multi-language VM, Marr and D'Hondt selected four questions~\cite{Marr2012}:
+Marr and D'Hondt~\cite{Marr2012} surveyed a variety of concurrent and parallel programming concepts to identify a unifying concurrency abstraction that could be implemented at the virtual machine level.\footnote{A 2009 survey by Marr, Haupt, and D'Hondt~\cite{Marr2009} found that virtual machines at the time did not provide intrinsic support for more than one specific model.} A unifying abstraction would enable each concurrency model in a high-level language to be represented in machine code without sacrificing performance or semantics.
+To identify concepts relevant for a multi-language virtual machine, Marr and D'Hondt selected four questions:
 
 \begin{enumerate}
-  \item Can the concept be reasonably implemented as a library? (A library implementation may suffer from loss of performance or semantics to an extent that the concept would warrant VM inclusion.)
+  \item Can the concept be reasonably implemented as a library?
+(A library implementation may suffer from loss of performance or semantics to an extent that the concept would warrant virtual machine inclusion.)
 
-  \item Does the concept require runtime support to guarantee its semantics? (Semantic guarantees may be enforced by a compiler but not the VM.)
+  \item Does the concept require runtime support to guarantee its semantics?
+(Semantic guarantees may be enforced by a compiler but not the virtual machine.
+For instance, immutable objects in one language may not be immutable in machine code, and therefore may be mutated by another language.)
 
-  \item Would runtime support significantly benefit performance? (Deeper knowledge of the concept may enable the VM to better optimize.)
+  \item Would runtime support benefit performance?
+(Deeper knowledge of the concept may enable the virtual machine to better optimize.
+Some concepts, though, may require knowledge of the underlying hardware to realize any performance improvements; such concepts would not necessarily benefit from runtime support.)
 
-  \item Is the concept already supported by a common VM like the Java Virtual Machine (JVM) or the Common Language Runtime (CLR)?
+  \item Is the concept already supported by a common virtual machine like the Java Virtual Machine (JVM) or the Common Language Runtime (CLR)?
 \end{enumerate}
@@
 From those questions, 26 concepts were identified that would benefit from better low-level support.
+Of those, 18 suffered from loss of semantic guarantees when compiled to machine code.
+With that in mind, Marr and D'Hondt derived the following requirements for a unifying abstraction:
+
+\begin{itemize}
+  \item Managed mutation: Concurrency models impose restrictions on how state can be modified; therefore, mutation must be able to be flexibly managed.
+
+  \item Managed execution: Similarly, invocation of operations on objects is usually restricted and must be able to be flexibly managed.
+
+  \item Ownership: Some owning entity usually regulates mutation and execution, so ownership should be supported in a way that allows for adaptable mutation and execution rules.
+
+  \item Leveled reflection: Reflection allows a program to examine its own structure and behavior at runtime.
+To guarantee safety, there needs to be a distinction between language-level reflection (restricted reflection operations supported by the high-level language) and meta-level reflection (unrestricted reflection operations on the low-level abstraction).
+
+  \item Enforceability: All defined restrictions should be enforceable on different concurrency models.
+Additionally, the semantics of the language in which a model is defined must be enforced regardless of where the model is used.
+\end{itemize}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=224pt]{obmop}
+  \caption{Marr and D'Hondt's object-based meta-object protocol~\cite{Marr2012}.}
+\label{figure:obmop}
+\end{figure}
+
+Given those requirements, Marr and D'Hondt defined an ownership-based meta-object protocol (MOP) that could describe a low-level abstraction, illustrated in Figure~\ref{figure:obmop}.
+The owner of an object, referred to as the domain, manages operations on all of the objects that it owns (namely, reading and writing of object fields and invocation of methods on objects).
+This satisfies the ownership requirement.
+Read and write operations are directed to the \code{readField} and \code{writeField} operations on the domain, satisfying the managed mutation requirement.
+Similarly, all method invocations are directed to the \code{exec}, satisfying the managed execution requirement.
+Threads are executed in a domain and specify whether execution occurs at the language level (with restricted reflection) or the meta level (with unrestricted reflection), satisfying the leveled reflection requirement.
+Globally shared resources are handled by the domain (specifically, the \code{readGlobal} and \code{writeGlobal} methods) if they might break semantics.
+Additionally, the \code{primitive*} operations allow callers to override the semantics of virtual machine primitives, completing the enforcability requirement.
+
+To demonstrate the suitability of the MOP as a low-level abstraction, Marr and D'Hondt implemented several concurrency models.
+The MOP was able to support LRSTM (a STM for the Smalltalk language), Clojure's agents (which rely on communicating threads), active objects (a variation on proxies), and several other models.
+While the MOP was able to successfully enforce the semantics of each model, the resulting performance cost suggested that the MOP must be implemented at the virtual machine level to sufficiently reduce the abstraction overhead.
+Marr and D'Hondt also note that regarding the owner of an object as the only entity able to restrict interaction is somewhat limiting.
+For instance, the behavior of the system when different semantics interact needs to be explicitly specified.
 

### 2016-10-27 paper/paper.tex
Wrote composability conflicts section

@@
 
+\section{Composability challenges}
+
+While concurrency models guarantee their own correctness, issues may arise when different models are used together.
+For example, an implementation of a STM that guarantees linearizability assumes that all shared resources are managed by the STM~\cite{Shavit1995}.
+However, if a thread communicates with other threads (in CSP, for instance), that assumption could lead to unexpected interleaving of operations~\cite{Swalens2014}.
+
+Swalens et al.~\cite{Swalens2014}, in a through study of concurrency models provided by Clojure, examined issues that arise when different models are combined.
+Two models are said to be composable if combining them does not introduce correctness issues.
+
+\subsection{Correctness criteria}
+
+In the study, safety and liveness were used as the two criteria for evaluating the correctness of a combination of models.
+Together, the two have historically been used to prove the correctness of distributed systems~\cite{Lamport1977}.
+
+Informally, safety guarantees that ``nothing bad will happen.'' In other words, given a correct input, a program will not produce an incorrect result~\cite{Swalens2014}.
+This property is also referred to as partial correctness.
+Safety is generally achieved in the management of shared resources.
+STM, for instance, only allows shared memory to be accessed through transactions; communicating threads only allow data to be shared through message passing.
+
+Liveness guarantees that ``something good will eventaully happen,'' or that a program will eventually terminate if its input is correct.
+Taken together, safety and liveness are referred to as total correctness---given a correct input, a program will terminate with the correct output~\cite{Swalens2014}.
+Deadlocks and livelocks are the primary obstacle to liveness and can arise when models are combined.
+
+\subsection{Possible conflicts}
+
+To discover the types of conflicts that might arise when combining different models, the study examined all pairwise combinations of Clojure's models.
+This process uncovered the following types of conflicts:
+
+\begin{itemize}
+  \item A model reexecutes code containing another concurrency model that performs an irrevocable action. (A STM transaction contains a proxy that performs a network request.)
+
+  \item A model reexecutes code that causes the reexecution to continually happen.
+(An STM transaction that consistently conflicts with another may cause a livelock, though Swalens et al.~note that Clojure dynamically prevents this type of conflict.)
+
+  \item A model that supports blocking operations is used within a model that does not expect blocking operations.
+(A deadlock might be caused when a read operation occurs within mutually recursive futures.)
+
+  \item A model does not guarantee safety or liveness by design, making composability impossible.
+(A model that does not guarantee safety would violate the safety guarantee of an outer model unless such a conflict was explicitly prevented.)
+\end{itemize}
+
 \section{Proposed abstractions}
 
-\subsection{Asynchronous events}
+%\subsection{Asynchronous events}
 
-Ziarek et al.~\cite{Ziarek2011} introduce asynchronous events as a possible abstraction. Their examples are heavily tied to Concurrent ML and might be difficult to integrate.
+% Ziarek et al.~\cite{Ziarek2011} introduce asynchronous events as a possible abstraction. Their examples are heavily tied to Concurrent ML and might be difficult to integrate.
 

### 2016-11-12 slides/slides.tex
Hammered out presentation intro

@@
 
+\usepackage{graphicx}
+\usepackage{tikz}
+
 % this actually isn't totally hideous:
-\usetheme{metropolis}
-\setsansfont{Noto Sans}
+\usetheme[numbering=counter, progressbar=foot, sectionpage=none]{metropolis}
+
+\graphicspath{{../figures/}}
 
@@
 \author{Dan Stelljes}
-\institute{University of Minnesota Morris}
 
@@
 
-  \begin{frame}{Concurrency Models}
-    Example thing goes here.
+  \section{Introduction}
+
+  \begin{frame}
+    \begin{figure}
+      \begin{tikzpicture}
+        \onslide<1->{
+          \node at (0,0) {
+            \includegraphics[width=300pt]{browser}
+          };
+        }
+
+        \onslide<2>{
+          \path (-2.25,2.75) [draw=none, fill=darkgray, opacity=0.75] ellipse (60pt and 25pt);
+          \node [text=white] at (-2.25,2.75) {
+            Multiple tabs
+          };
+
+          \draw (2.25,2.25) [draw=none, fill=darkgray, opacity=0.75] ellipse (60pt and 25pt);
+          \node [text=white] at (2.25,2.25) {
+            Suggestions
+          };
+
+          \draw (0,0.5) [draw=none, fill=darkgray, opacity=0.75] ellipse (60pt and 25pt);
+          \node [text=white] at (0,0.5) {
+            Page rendering
+          };
+
+          \draw (-3,-1.5) [draw=none, fill=darkgray, opacity=0.75] ellipse (60pt and 25pt);
+          \node [text=white] at (-3,-1.5) {
+            Event handling
+          };
+
+          \draw (2.5,-1.5) [draw=none, fill=darkgray, opacity=0.75] ellipse (60pt and 25pt);
+          \node [text=white] at (2.5,-1.5) {
+            Background processes
+          }; 
+        }
+
+        \onslide<3>{
+          \node at (0,-0.62) {
+            \includegraphics[width=220pt]{goals}
+          };
+        }
+      \end{tikzpicture}
+    \end{figure}
   \end{frame}

### 2016-11-14 slides/slides.tex
Wrote background slides

@@
 
+\usepackage[USenglish]{babel}
+\usepackage[backend=bibtex,citestyle=authortitle]{biblatex}
 \usepackage{graphicx}
@@
 \author{Dan Stelljes}
+
+\bibliography{references}
 
@@
   \end{frame}
+
+  \section{Background}
+
+  \begin{frame}
+    \frametitle{Concurrency}
+    \framesubtitle{Intuitive definition}
+
+    \onslide<1>{
+      \begin{columns}[T]
+        \begin{column}{0.48\textwidth}
+          \centering
+          sequential
+        \end{column}
+
+        \hfill
+
+        \begin{column}{0.48\textwidth}
+          \centering
+          concurrent
+        \end{column}
+      \end{columns}
+    }
+
+    \onslide<2>{
+      \begin{columns}[T]
+        \begin{column}{0.48\textwidth}
+          \centering
+          concurrent
+        \end{column}
+
+        \hfill
+
+        \begin{column}{0.48\textwidth}
+          \centering
+          parallel
+        \end{column}
+      \end{columns}
+    }
+  \end{frame}
+
+  \begin{frame}
+    \frametitle{Concurrency}
+    \framesubtitle{Formal definition}
+
+    \textbf{The ``happens before'' ($\rightarrow$) relation~\footcite{Lamport1977}}
+
+    $A \rightarrow B$ if one of the following is true:
+
+    \begin{enumerate}
+      \item $A$ and $B$ are operations in the same thread and $A$ occurs before $B$.
+      \item $A$ is the sending of a message by one thread and $B$ is the receipt of the same message by another thread.
+    \end{enumerate}
+
+    $A$ and $B$ are said to be concurrent if $A \nrightarrow B$ and $B \nrightarrow A$.
+  \end{frame}
+
+  \begin{frame}
+    \frametitle{Consistency models}
+
+    \begin{itemize}
+      \item \textbf{Sequential:} Does the history of the program yield a correct result?
+      \item \textbf{Concurrent:} Does every possible history of the program yield a correct result?
+    \end{itemize}
+  \end{frame}
+
+  \begin{frame}
+    \frametitle{Complications}
+
+    \onslide<1->{
+      \centering
+      \includegraphics{single-thread-read-write}
+    }
+
+    \onslide<2->{
+      \centering
+      \includegraphics{multiple-threads-read-write}
+    }
+  \end{frame}
+
+  \begin{frame}
+    \frametitle{Complications}
+
+    \begin{figure}
+      \centering
+      \includegraphics{noninstantaneity}
+    \end{figure}
+
+    \begin{itemize}
+      \item Nothing is truly instantaneous.
+      \item The history of a program may be influenced by the time it takes for operations to complete.
+    \end{itemize}
+  \end{frame}
+
+  \begin{frame}
+    \frametitle{Linearizability}
+
+    \begin{figure}
+      \centering
+      \includegraphics{linearizability}
+    \end{figure}
+
+    \begin{itemize}
+      \item Linearizability guarantees that the completion of an operation on a single object will appear to be instantaneous.
+      \item The results of a linearizable operation will be visible as soon as the operation is complete.
+    \end{itemize}
+  \end{frame}
+
+  \begin{frame}
+    \frametitle{Serializability}
+
+    \begin{figure}
+      \centering
+      \includegraphics{serializability}
+    \end{figure}
+
+    \begin{itemize}
+      \item Serializability guarantees that operations can occur in any order as long as an equivalent serial history exists.
+      \item While a serializable set of operations is being executed, it appears to be the only set of operations being executed.
+    \end{itemize}
+  \end{frame}
+
+  \begin{frame}
+    \frametitle{Strict serializability}
+
+    \begin{itemize}
+      \item Linearizability is a guarantee of consistency.
+      \item Serializability is a guarantee of isolation.
+      \item Together, linearizability and serializability guarantee strict serializability---a history is equivalent to some serial execution order and that order corresponds to the execution order in real time.
+    \end{itemize}
+  \end{frame}
+
 \end{document}

### 2016-11-14 slides/slides.tex
Scaffolded the rest of the slide deck

@@
 \usepackage[backend=bibtex,citestyle=authortitle]{biblatex}
+\usepackage[scale=2]{ccicons}
 \usepackage{graphicx}
@@
 % this actually isn't totally hideous:
-\usetheme[numbering=counter, progressbar=foot, sectionpage=none]{metropolis}
+\usetheme[numbering=counter, progressbar=frametitle, sectionpage=none]{metropolis}
 
@@
 
+  \subsection{Concurrency}
+
   \begin{frame}
@@
 
+  \subsection{Consistency}
+
   \begin{frame}
@@
     \end{figure}
+  \end{frame}
 
-    \begin{itemize}
-      \item Nothing is truly instantaneous.
-      \item The history of a program may be influenced by the time it takes for operations to complete.
-    \end{itemize}
-  \end{frame}
+  \subsection{Correctness}
 
@@
 
+  \section{Common concurrency models}
+
+  \begin{frame}
+    \frametitle{Concurrency models}
+
+    \begin{itemize}
+      \item Modern languages usually support several different models.
+      \item Common concurrency models can be generalized:~\footcite{Swalens2014}
+      \begin{enumerate}
+        \item Atomic variables
+        \item Software transactional memory (STM)
+        \item Communicating threads
+        \item Proxies
+      \end{enumerate}
+    \end{itemize}
+  \end{frame}
+
+  \subsection{Atomic variables}
+
+  \begin{frame}
+  \end{frame}
+
+  \subsection{Software transactional memory}
+
+  \begin{frame}
+  \end{frame}
+
+  \subsection{Communicating threads}
+
+  \begin{frame}
+  \end{frame}
+
+  \subsection{Proxies}
+
+  \begin{frame}
+  \end{frame}
+
+  \section{Composability challenges}
+
+  \subsection{Possible conflicts}
+
+  \section{Proposed abstractions}
+
+  \subsection{Ownership-based meta-object protocol}
+
+  \section{Conclusion}
+
+  \begin{frame}
+    \frametitle{Future work}
+  \end{frame}
+
+  \begin{frame}[standout]
+    \centering
+
+    \url{github.com/dstelljes/senior-sem}
+
+    \vfill
+
+    \ccbyncsa
+  \end{frame}
+
 \end{document}

### 2016-11-15 slides/slides.tex
Added compare-and-swap; we'll see if it lasts
Hastily filled in rest of slide deck

@@
 \usepackage{graphicx}
+\usepackage{listings}
 \usepackage{tikz}
@@
       \item Serializability is a guarantee of isolation.
-      \item Together, linearizability and serializability guarantee strict serializability---a history is equivalent to some serial execution order and that order corresponds to the execution order in real time.
+      \item Together, linearizability and serializability guarantee strict serializability---a history is equivalent to some serial execution order and that order corresponds to the execution order in real time.~\footcite{Herlihy1990}
     \end{itemize}
@@
       \item Modern languages usually support several different models.
-      \item Common concurrency models can be generalized:~\footcite{Swalens2014}
+      \item Common concurrency models fall into four general categories:~\footcite{Swalens2014}
       \begin{enumerate}
@@
 
+  \begin{frame}[fragile]
+    \frametitle{Atomic variables}
+
+    \begin{lstlisting}[basicstyle=\scriptsize\ttfamily,language=C]
+      void unsafe_increment() {
+        *variable++;
+      }
+    \end{lstlisting}
+% "good indentation? lol get bent" —Beamer
+\end{frame}
+
+  \begin{frame}[fragile]
+    \frametitle{Atomic variables}
+
+    \begin{lstlisting}[basicstyle=\scriptsize\ttfamily,language=C]
+      int fetch_and_add(int* variable, int increment) {
+        // obtain exclusive control of variable
+
+        int value = *variable + increment;
+        *variable = value;
+
+        // release exclusive control of variable
+
+        return value;
+      }
+
+      void safe_increment() {
+        fetch_and_add(variable, 1);
+      }
+    \end{lstlisting}
+\end{frame}
+
+  \begin{frame}[fragile]
+    \frametitle{Atomic variables}
+
+    \begin{lstlisting}[basicstyle=\scriptsize\ttfamily,language=Java]
+      AtomicInteger variable = new AtomicInteger(23);
+      variable.incrementAndGet();
+    \end{lstlisting}
+
+    \vfill
+
+    \begin{lstlisting}[basicstyle=\scriptsize\ttfamily,language=Lisp]
+      (def variable (atom 23))
+      (swap! variable inc)
+    \end{lstlisting}
+\end{frame}
+
   \begin{frame}
+    \frametitle{Atomic variables}
+
+    Atomic variables are\ldots
+
+    \begin{itemize}
+      \item used to share objects that are shared by multiple threads but don't require coordinated updates
+      \item usually implemented in hardware, guaranteeing linearizability at the lowest level possible
+      \item used to implement higher-level concurrency control mechanisms and concurrent data structures
+    \end{itemize}
   \end{frame}
@@
   \begin{frame}
+    \frametitle{Software transactional memory}
+
+    Another approach to working with shared memory:~\footcite{Shavit1995}
+
+    \begin{enumerate}
+      \item A thread writes to a shared memory location, keeping track of the transaction in a log.
+      \item If there are conflicting changes at the end of the transaction, the transaction is aborted and retried until it eventually succeeds.
+      \item Once a transaction succeeds, changes will become visible to other threads.
+    \end{enumerate}
+  \end{frame}
+
+  \begin{frame}[fragile]
+    \frametitle{Software transactional memory}
+
+    \begin{lstlisting}[basicstyle=\scriptsize\ttfamily,language=Java]
+
+      import akka.stm.*;
+
+      Ref<Integer> reference = new Ref<Integer>(23);
+
+      new Atomic<Integer>() {
+        public Integer atomically() {
+          int value = reference.get();
+          reference.set(value++);
+
+          return value;
+        }
+      }.execute();
+    \end{lstlisting}
+
+    \vfill
+
+    \begin{lstlisting}[basicstyle=\scriptsize\ttfamily,language=Lisp]
+      (def reference (ref 23))
+      (dosync (alter reference inc))
+    \end{lstlisting}
+\end{frame}
+
+  \begin{frame}
+    \frametitle{Software transactional memory}
+
+    STM is\ldots
+
+    \begin{itemize}
+      \item used to share objects that need to be accessed and modified by multiple threads
+      \item a simple alternative to lock-based programming
+      \item susceptible to performance decreases caused by maintaining a transaction log and retrying transactions
+      \item unable to execute operations that cannot be undone or retried
+    \end{itemize}
   \end{frame}
@@
   \begin{frame}
+    \frametitle{Communicating threads}
+
+    \begin{itemize}
+      \item Communicating sequential processes (CSP) describes systems as independent processes that communicate synchronously through predefined channels.~\footcite{Hoare1978}
+      \item The actor model relies on asynchronous message passing to specific entities.~\footcite{Agha1986}
+      \item The agent model emulates atomic references by containing a shared object in a separate process.~\footcite{Swalens2014}
+    \end{itemize}
+  \end{frame}
+
+  \begin{frame}
+    \frametitle{Communicating threads}
+  \end{frame}
+
+  \begin{frame}
+    \frametitle{Communicating threads}
+
+    Communicating threads are\ldots
+
+    \begin{itemize}
+      \item used to avoid working with shared objects
+      \item helpful when communicating with external systems or running an event loop
+    \end{itemize}
   \end{frame}
@@
   \begin{frame}
+    \frametitle{Proxies}
+
+    Proxies serve as placeholders for results of concurrent operations.
+
+    \begin{itemize}
+      \item \textbf{Futures:} eventually resolved to the result of the completed operation; result is accessed implicitly
+      \item \textbf{Promises:} exist as independent objects; result is accessed explicitly
+    \end{itemize}
+  \end{frame}
+
+  \begin{frame}
+    \frametitle{Proxies}
+  \end{frame}
+
+  \begin{frame}
+    \frametitle{Proxies}
+
+    Proxies are\ldots
+
+    \begin{itemize}
+      \item used to execute long-running operations without blocking other operations
+      \item easily passed around and composed
+    \end{itemize}
   \end{frame}
@@
 
-  \subsection{Possible conflicts}
+  \begin{frame}
+    \frametitle{Composability challenges}
+    Need some stuff here.~\footcite{Swalens2014}
+  \end{frame}
+
+  \begin{frame}
+    \frametitle{Correctness criteria}
+
+    \begin{itemize}
+      \item \textbf{Safety:} nothing bad will happen (partial correctness)
+      \item \textbf{Liveness:} something good will eventually happen (with safety, total correctness)
+    \end{itemize}
+  \end{frame}
+
+  \begin{frame}
+    \frametitle{Possible conflicts}
+
+    \begin{itemize}
+      \item A model reexecutes code containing another model that performs an irrevocable action.
+      \item A model reexecutes code that causes the reexecution to continually happen.
+      \item A model that supports blocking operations is used within a model that does not support blocking operations.
+      \item A model does not guarantee safety or liveness by design, making composability impossible.
+    \end{itemize}
+  \end{frame}
 
@@
   \subsection{Ownership-based meta-object protocol}
+
+  \begin{frame}
+    \frametitle{Ownership-based meta-object protocol}
+
+    \begin{enumerate}
+      \item Can the concept be reasonably implemented as a library?
+      \item Does the concept require runtime support to guarantee its semantics?
+      \item Would runtime support benefit performance?
+      \item Is the concept already supported by a common virtual machine (e.g., the JVM or the CLR)?
+    \end{enumerate}
+  \end{frame}
+
+  \begin{frame}
+    \frametitle{Ownership-based meta-object protocol}
+  \end{frame}
+
+  \begin{frame}
+    \frametitle{Ownership-based meta-object protocol}
+
+    \centering
+
+    \includegraphics[width=150pt]{obmop}
+  \end{frame}
 
@@
 
+    Thanks to Elena Machkasova, K.K. Lamberty, and Matthew Justin for their guidance and suggestions.
+
+    \vfill
+
     \url{github.com/dstelljes/senior-sem}
@@
 
-    \ccbyncsa
+    \ccbyncsa{}
   \end{frame}

### 2016-11-16 slides/slides.tex
Cleaned up presentation intro/background and typeset diagrams

@@
 
+\usepackage{appendixnumberbeamer}
 \usepackage[USenglish]{babel}
@@
 \usetheme[numbering=counter, progressbar=frametitle, sectionpage=none]{metropolis}
+\setbeamertemplate{bibliography item}[text]
+\setbeamertemplate{frametitle continuation}{}
 
@@
 
+\AtEveryBibitem{\clearfield{note}}
+\AtEveryBibitem{\clearfield{url}}
+
 \begin{document}
@@
   \begin{frame}
-    \begin{figure}
-      \begin{tikzpicture}
-        \onslide<1->{
-          \node at (0,0) {
-            \includegraphics[width=300pt]{browser}
-          };
-        }
-
-        \onslide<2>{
-          \path (-2.25,2.75) [draw=none, fill=darkgray, opacity=0.75] ellipse (60pt and 25pt);
-          \node [text=white] at (-2.25,2.75) {
-            Multiple tabs
-          };
-
-          \draw (2.25,2.25) [draw=none, fill=darkgray, opacity=0.75] ellipse (60pt and 25pt);
-          \node [text=white] at (2.25,2.25) {
-            Suggestions
-          };
-
-          \draw (0,0.5) [draw=none, fill=darkgray, opacity=0.75] ellipse (60pt and 25pt);
-          \node [text=white] at (0,0.5) {
-            Page rendering
-          };
-
-          \draw (-3,-1.5) [draw=none, fill=darkgray, opacity=0.75] ellipse (60pt and 25pt);
-          \node [text=white] at (-3,-1.5) {
-            Event handling
-          };
-
-          \draw (2.5,-1.5) [draw=none, fill=darkgray, opacity=0.75] ellipse (60pt and 25pt);
-          \node [text=white] at (2.5,-1.5) {
-            Background processes
-          };
-        }
-
-        \onslide<3>{
-          \node at (0,-0.62) {
-            \includegraphics[width=220pt]{goals}
-          };
-        }
-      \end{tikzpicture}
-    \end{figure}
+    \centering
+    \resizebox{\linewidth}{!}{\input{../figures/introduction.tikz}}
   \end{frame}
@@
     \frametitle{Concurrency}
-    \framesubtitle{Intuitive definition}
-
-    \onslide<1>{
-      \begin{columns}[T]
-        \begin{column}{0.48\textwidth}
-          \centering
-          sequential
-        \end{column}
-
-        \hfill
-
-        \begin{column}{0.48\textwidth}
-          \centering
-          concurrent
-        \end{column}
-      \end{columns}
-    }
-
-    \onslide<2>{
-      \begin{columns}[T]
-        \begin{column}{0.48\textwidth}
-          \centering
-          concurrent
-        \end{column}
-
-        \hfill
-
-        \begin{column}{0.48\textwidth}
-          \centering
-          parallel
-        \end{column}
-      \end{columns}
-    }
-  \end{frame}
-
-  \begin{frame}
-    \frametitle{Concurrency}
-    \framesubtitle{Formal definition}
 
@@
 
+    If $A \rightarrow B$ and $B \rightarrow C$, then $A \rightarrow C$.
+
     $A$ and $B$ are said to be concurrent if $A \nrightarrow B$ and $B \nrightarrow A$.
@@
 
-  \subsection{Consistency}
+  \subsection{Complications}
 
   \begin{frame}
-    \frametitle{Consistency models}
+    \frametitle{Complications}
 
     \begin{itemize}
-      \item \textbf{Sequential:} Does the history of the program yield a correct result?
-      \item \textbf{Concurrent:} Does every possible history of the program yield a correct result?
+      \item \textbf{Sequential program:} Does the order of operations yield a correct result?
+      \vfill
+      \item \textbf{Concurrent program:} Does \emph{every possible} order of operations yield a correct result?
     \end{itemize}
@@
     \frametitle{Complications}
+    \centering
 
-    \onslide<1->{
+    \textbf{Single thread:}
+    \vfill
+    \resizebox{0.9\linewidth}{!}{\input{../figures/single-thread-read-write.tikz}}
+  \end{frame}
+
+  \begin{frame}
+    \frametitle{Complications}
     \centering
-      \includegraphics{single-thread-read-write}
-    }
 
-    \onslide<2->{
-      \centering
-      \includegraphics{multiple-threads-read-write}
-    }
+    \textbf{Multiple threads:}
+    \vfill
+    \resizebox{0.9\linewidth}{!}{\input{../figures/multiple-threads-read-write.tikz}}
   \end{frame}
@@
 
-    \begin{figure}
     \centering
-      \includegraphics{noninstantaneity}
-    \end{figure}
+    \input{../figures/noninstantaneity.tikz}
   \end{frame}
 
-  \subsection{Correctness}
+  \subsection{Consistency models}
 
@@
       \centering
-      \includegraphics{linearizability}
+      \input{../figures/linearizability.tikz}
     \end{figure}
+
+    \vfill
 
@@
       \centering
-      \includegraphics{serializability}
+      \input{../figures/serializability.tikz}
     \end{figure}
 
+    \vfill
+
     \begin{itemize}
-      \item Serializability guarantees that operations can occur in any order as long as an equivalent serial history exists.
+      \item Serializability guarantees that operations can occur in any order as long as an equivalent sequential ordering exists.
       \item While a serializable set of operations is being executed, it appears to be the only set of operations being executed.
@@
       \item Serializability is a guarantee of isolation.
-      \item Together, linearizability and serializability guarantee strict serializability---a history is equivalent to some serial execution order and that order corresponds to the execution order in real time.~\footcite{Herlihy1990}
+      \item Together, linearizability and serializability guarantee strict serializability---an ordering of operations is equivalent to some sequential ordering and that ordering corresponds to the execution ordering in real time.~\footcite{Herlihy1990}
     \end{itemize}
@@
     \centering
-
     Thanks to Elena Machkasova, K.K. Lamberty, and Matthew Justin for their guidance and suggestions.
-
     \vfill
-
-    \url{github.com/dstelljes/senior-sem}
-
+    \href{https://github.com/dstelljes/senior-sem}{github.com/dstelljes/senior-sem}
     \vfill
-
     \ccbyncsa{}
@@
 
+  \appendix{}
+
+  \begin{frame}[allowframebreaks]
+    \frametitle{References}
+
+    \printbibliography{}
+  \end{frame}
+
 \end{document}

### 2016-11-17 slides/slides.tex
Replaced examples with more crappy drawings

--- 2016-11-14 slides/slides.tex
@@
-\documentclass{beamer}
+\documentclass[aspectratio=1610]{beamer}
 
+\usepackage{appendixnumberbeamer}
 \usepackage[USenglish]{babel}
@@
 \usetheme[numbering=counter, progressbar=frametitle, sectionpage=none]{metropolis}
-
-\graphicspath{{../figures/}}
+\setbeamertemplate{bibliography item}[text]
+\setbeamertemplate{frametitle continuation}{}
 
@@
   \begin{frame}
-    \begin{figure}
-      \begin{tikzpicture}
-        \onslide<1->{
-          \node at (0,0) {
-            \includegraphics[width=300pt]{browser}
-          };
-        }
-
-        \onslide<2>{
-          \path (-2.25,2.75) [draw=none, fill=darkgray, opacity=0.75] ellipse (60pt and 25pt);
-          \node [text=white] at (-2.25,2.75) {
-            Multiple tabs
-          };
-
-          \draw (2.25,2.25) [draw=none, fill=darkgray, opacity=0.75] ellipse (60pt and 25pt);
-          \node [text=white] at (2.25,2.25) {
-            Suggestions
-          };
-
-          \draw (0,0.5) [draw=none, fill=darkgray, opacity=0.75] ellipse (60pt and 25pt);
-          \node [text=white] at (0,0.5) {
-            Page rendering
-          };
-
-          \draw (-3,-1.5) [draw=none, fill=darkgray, opacity=0.75] ellipse (60pt and 25pt);
-          \node [text=white] at (-3,-1.5) {
-            Event handling
-          };
-
-          \draw (2.5,-1.5) [draw=none, fill=darkgray, opacity=0.75] ellipse (60pt and 25pt);
-          \node [text=white] at (2.5,-1.5) {
-            Background processes
-          }; 
-        }
-
-        \onslide<3>{
-          \node at (0,-0.62) {
-            \includegraphics[width=220pt]{goals}
-          };
-        }
-      \end{tikzpicture}
-    \end{figure}
+    \centering
+    \resizebox{0.8\linewidth}{!}{\input{../figures/introduction.tikz}}
   \end{frame}
@@
 
-  \subsection{Consistency}
+  \subsection{Complications}
 
   \begin{frame}
-    \frametitle{Consistency models}
+    \frametitle{Complications}
 
     \begin{itemize}
-      \item \textbf{Sequential:} Does the history of the program yield a correct result?
-      \item \textbf{Concurrent:} Does every possible history of the program yield a correct result?
+      \onslide<1->{
+        \item \textbf{Sequential program:} Does the order of operations yield a correct result?
+      }
+
+      \vfill
+
+      \onslide<2->{
+        \item \textbf{Concurrent program:} Does \emph{every possible} order of operations yield a correct result?
+      }
     \end{itemize}
@@
     \begin{itemize}
-      \item Linearizability is a guarantee of consistency.
-      \item Serializability is a guarantee of isolation.
-      \item Together, linearizability and serializability guarantee strict serializability---a history is equivalent to some serial execution order and that order corresponds to the execution order in real time.
+      \item Linearizability \emph{and} serializability yield strict serializability, which guarantees both consistency and isolation.
     \end{itemize}
+
+    \vfill
+
+    \textbf{Strict serializability~\footcite{Herlihy1990}}
+
+    An ordering of operations is equivalent to some sequential ordering and that ordering corresponds to the order of execution in real time.
   \end{frame}
 
-  \section{Common concurrency models}
-
-  \begin{frame}
-    \frametitle{Concurrency models}
-
-    \begin{itemize}
-      \item Modern languages usually support several different models.
-      \item Common concurrency models can be generalized:~\footcite{Swalens2014}
-      \begin{enumerate}
-        \item Atomic variables
-        \item Software transactional memory (STM)
-        \item Communicating threads
-        \item Proxies
-      \end{enumerate}
-    \end{itemize}
-  \end{frame}
+  \section{Concurrency models}
 
@@
   \begin{frame}
+    \frametitle{Atomic variables}
+    \centering
+
+    \includegraphics[width=220pt]{../figures/nonatomic}
   \end{frame}
 
-  \subsection{Software transactional memory}
+  \begin{frame}
+    \frametitle{Atomic variables}
+    \centering
+
+    \includegraphics[width=220pt]{../figures/atomic}
+  \end{frame}
 
   \begin{frame}
+    \frametitle{Atomic variables}
+    \centering
+
+    \includegraphics[width=220pt]{../figures/lock}
+  \end{frame}
+
+  \begin{frame}
+    \frametitle{Atomic variables}
+    \centering
+
+    \includegraphics[width=220pt]{../figures/multiple-locks}
+  \end{frame}
+
+  \subsection{Transactional memory}
+
+  \begin{frame}
+    \frametitle{Transactional memory}
+    \centering
+
+    \includegraphics[width=220pt]{../figures/stm}
+  \end{frame}
+
+  \begin{frame}
+    \frametitle{Transactional memory}
+
+    Software transactional memory (STM) is an optimistic approach to working with shared memory:~\footcite{Shavit1995}
+
+    \begin{enumerate}
+      \item A thread writes to a shared memory location, keeping track of the transaction in a log.
+      \item If there are conflicting changes at the end of the transaction, the transaction is aborted and retried.
+      \item If there are no conflicts, the changes are committed and become visible.
+    \end{enumerate}
   \end{frame}
@@
   \begin{frame}
+    \frametitle{Communicating threads}
+    \centering
+
+    \includegraphics[width=220pt]{../figures/agent}
+    \vfill
+    \textbf{The agent model:} An isolated thread handles all operations on an object.~\footcite{Swalens2014}
   \end{frame}
 
-  \subsection{Proxies}
+  \begin{frame}
+    \frametitle{Communicating threads}
+    \centering
+
+    \includegraphics[width=220pt]{../figures/csp}
+    \vfill
+    \textbf{Communicating sequential processes:} Independent threads communicate synchronously through predefined channels.~\footcite{Hoare1978}
+  \end{frame}
 
   \begin{frame}
+    \frametitle{Communicating threads}
+    \centering
+
+    \includegraphics[width=220pt]{../figures/actor}
+    \vfill
+    \textbf{The actor model:} Independent threads send messages to known addresses.~\footcite{Agha1986}
   \end{frame}
@@
 
-  \subsection{Possible conflicts}
+  \begin{frame}
+    \frametitle{Composability criteria}
+
+    \textbf{How do we know that models are composable?}
+
+    \begin{itemize}
+      \item Safety: ``Nothing bad will happen!'' (The output of a program or algorithm will not be incorrect.)
+      \item Liveness: ``Something will eventually happen!'' (The program or algorithm will terminate.)
+    \end{itemize}
+
+    Two models are composable if using them within each other doesn't compromise safety or liveness.~\footcite{Swalens2014}
+  \end{frame}
+
+  \begin{frame}
+    \frametitle{Composability criteria}
+    \centering
+
+    \includegraphics[width=360pt]{../figures/clojure-survey}
+  \end{frame}
+
+  \begin{frame}
+    \frametitle{Possible conflicts}
+
+    \begin{itemize}
+      \onslide<1->{
+        \item A model reexecutes code containing another model that performs an irrevocable action.
+      }
+      \vfill
+      \onslide<2->{
+        \item A model reexecutes code that causes the reexecution to continually happen.
+      }
+      \vfill
+      \onslide<3->{
+        \item A model that supports blocking operations is used within a model that does not support blocking operations.
+      }
+      \vfill
+      \onslide<4->{
+        \item A model does not guarantee safety or liveness by design, making composability impossible.
+      }
+    \end{itemize}
+  \end{frame}
+
+  \begin{frame}
+    \frametitle{Conclusions}
+
+    \begin{itemize}
+      \onslide<1->{
+        \item The language may ensure that some bad interactions don't happen.
+      }
+      \vfill
+      \onslide<2->{
+        \item Sometimes, it's best to just use a different concurrency model.
+      }
+      \vfill
+      \onslide<3->{
+        \item Future work would identify common ``building blocks'' (thread creation, message passing, etc.) that could be composed to express existing concurrency models and combinations of models.
+      }
+    \end{itemize}
+  \end{frame}
 
@@
 
-  \section{Conclusion}
+  \begin{frame}
+    \frametitle{Proposed abstractions}
+
+    \textbf{What characteristics should a unified abstraction have?}~\footcite{Marr2012}
+
+    \begin{itemize}
+      \onslide<1->{
+        \item Ownership: All objects are owned by something, so ownership should be supported in a way that supports regulated access.
+      }
+      \vfill
+      \onslide<2->{
+        \item Managed mutation and execution: The owner of an object should determine how it be read and modified.
+      }
+      \vfill
+      \onslide<3->{
+        \item Leveled access: Threads should either have model-level access or abstraction-level access.
+      }
+      \vfill
+      \onslide<4->{
+        \item Enforceability: The model's semantics should be enforced regardless of where the model is used.
+      }
+    \end{itemize}
+  \end{frame}
 
   \begin{frame}
-    \frametitle{Future work}
+    \frametitle{Ownership-based meta-object protocol}
+
+    \centering
+
+    \includegraphics[width=250pt]{../figures/obmop-simple.png}
+  \end{frame}
+
+  \begin{frame}
+    \frametitle{Conclusions}
   \end{frame}
@@
     \centering
+    Thanks to Elena Machkasova, K.K. Lamberty, and Matthew Justin for their guidance and suggestions.
+    \vfill
+    \href{https://github.com/dstelljes/senior-sem}{github.com/dstelljes/senior-sem}
+    \vfill
+    \ccbyncsa{}
+  \end{frame}
 
-    \url{github.com/dstelljes/senior-sem}
+  \appendix{}
 
-    \vfill
+  \begin{frame}[allowframebreaks]
+    \frametitle{References}
 
-    \ccbyncsa
+    \printbibliography{}
   \end{frame}

### 2016-11-17 slides/slides.tex
Cleaned up some stuff pre-practice talk

@@
     \frametitle{Atomic variables}
+
+    \begin{columns}
+      \begin{column}{0.5\textwidth}
         \centering
+        \includegraphics[width=160pt]{../figures/nonatomic}
+      \end{column}
+      \hfill
+      \begin{column}{0.5\textwidth}
+        $x = 23$
 
-    \includegraphics[width=220pt]{../figures/nonatomic}
+        \begin{enumerate}
+          \item $A$ reads $x$
+          \item $B$ reads $x$
+          \item $A$ increments value
+          \item $A$ writes incremented value to $x$
+          \item $B$ increments value
+          \item $B$ writes incremented value to $x$
+        \end{enumerate}
+
+        $x = 24$
+      \end{column}
+    \end{columns}
   \end{frame}
@@
     \frametitle{Atomic variables}
+
+    \begin{columns}
+      \begin{column}{0.5\textwidth}
         \centering
+        \includegraphics[width=160pt]{../figures/atomic}
+      \end{column}
+      \hfill
+      \begin{column}{0.5\textwidth}
+        $x = 23$
 
-    \includegraphics[width=220pt]{../figures/atomic}
+        \begin{enumerate}
+          \item $A$ calls \textsc{fetch-and-increment} on $x$
+          \item $B$ calls \textsc{fetch-and-increment} on $x$
+        \end{enumerate}
+
+        $x = 25$
+      \end{column}
+    \end{columns}
   \end{frame}
@@
     \vfill
-    \textbf{The agent model:} An isolated thread handles all operations on an object.~\footcite{Swalens2014}
+    \textbf{Agents:} An isolated thread handles all operations on an object.~\footcite{Swalens2014}
   \end{frame}
@@
     \vfill
-    \textbf{Communicating sequential processes:} Independent threads communicate synchronously through predefined channels.~\footcite{Hoare1978}
+    \textbf{Communicating sequential processes (CSP):} Independent threads communicate synchronously through predefined channels.~\footcite{Hoare1978}
   \end{frame}
@@
 
-    \textbf{How do we know that models are composable?}
+    \textbf{How do we know that models are composable?~\footcite{Swalens2014}}
 
@@
 
-    Two models are composable if using them within each other doesn't compromise safety or liveness.~\footcite{Swalens2014}
+    Two models are composable if using them within each other doesn't compromise safety or liveness.
   \end{frame}
@@
 
-  \section{Proposed abstractions}
+  \section{Unifying abstractions}
 
@@
 
-    \textbf{What characteristics should a unified abstraction have?}~\footcite{Marr2012}
+    \textbf{What characteristics should a unifying abstraction have?}~\footcite{Marr2012}
 
@@
       \onslide<2->{
-        \item Managed mutation and execution: The owner of an object should determine how it be read and modified.
+        \item Managed mutation and execution: The owner of an object should determine how the object should be read and modified.
       }
@@
       \onslide<4->{
-        \item Enforceability: The model's semantics should be enforced regardless of where the model is used.
+        \item Enforceability: The abstraction should be able to enforce all model rules.
       }
@@
     \frametitle{Conclusions}
+
+    \begin{itemize}
+      \onslide<1->{
+        \item An abstraction like the OBMOP must be implemented at a low level to achieve acceptable performance.
+      }
+      \vfill
+      \onslide<2->{
+        \item The OBMOP has some limitations: The owner is the only entity that enforces rules, and the abstraction doesn't guarantee things like freedom from deadlock.
+      }
+      \vfill
+      \onslide<3->{
+        \item This approach has potential: The authors were able to implement STM, CSP, agents, and actors in less than 500 total lines of code.
+      }
+    \end{itemize}
   \end{frame}

### 2016-11-18 slides/slides.tex
Dropped OBMOP and restructured a little bit

@@
 \usepackage{listings}
+\usepackage{multicol}
 \usepackage{tikz}
@@
 
-  \section{Introduction}
-
   \begin{frame}
@@
     \resizebox{0.8\linewidth}{!}{\input{../figures/introduction.tikz}}
+  \end{frame}
+
+  \begin{frame}
+    \begin{multicols}{2}
+      \tableofcontents
+    \end{multicols}
   \end{frame}
@@
 
+  \addtocontents{toc}{\hfill}
+
   \section{Concurrency models}
@@
 
-  \subsection{Transactional memory}
+  \subsection{Software transactional memory}
 
   \begin{frame}
-    \frametitle{Transactional memory}
+    \frametitle{Software transactional memory}
     \centering
@@
   \begin{frame}
-    \frametitle{Transactional memory}
+    \frametitle{Software transactional memory}
 
@@
 
-  \section{Composability challenges}
+  % never change, LaTeX
+  \addtocontents{toc}{\newpage}
+
+  \section{Composability}
+
+  \subsection{Correctness criteria}
 
   \begin{frame}
-    \frametitle{Composability criteria}
+    \frametitle{Correctness criteria}
 
@@
   \end{frame}
+
+  \subsection{Possible conflicts}
 
@@
 
+  \subsection{Conclusions}
+
   \begin{frame}
@@
         \item Future work would identify common ``building blocks'' (thread creation, message passing, etc.) that could be composed to express existing concurrency models and combinations of models.
-      }
-    \end{itemize}
-  \end{frame}
-
-  \section{Unifying abstractions}
-
-  \subsection{Ownership-based meta-object protocol}
-
-  \begin{frame}
-    \frametitle{Proposed abstractions}
-
-    \textbf{What characteristics should a unifying abstraction have?}~\footcite{Marr2012}
-
-    \begin{itemize}
-      \onslide<1->{
-        \item Ownership: All objects are owned by something, so ownership should be supported in a way that supports regulated access.
-      }
-      \vfill
-      \onslide<2->{
-        \item Managed mutation and execution: The owner of an object should determine how the object should be read and modified.
-      }
-      \vfill
-      \onslide<3->{
-        \item Leveled access: Threads should either have model-level access or abstraction-level access.
-      }
-      \vfill
-      \onslide<4->{
-        \item Enforceability: The abstraction should be able to enforce all model rules.
-      }
-    \end{itemize}
-  \end{frame}
-
-  \begin{frame}
-    \frametitle{Ownership-based meta-object protocol}
-
-    \centering
-
-    \includegraphics[width=250pt]{../figures/obmop-simple.png}
-  \end{frame}
-
-  \begin{frame}
-    \frametitle{Conclusions}
-
-    \begin{itemize}
-      \onslide<1->{
-        \item An abstraction like the OBMOP must be implemented at a low level to achieve acceptable performance.
-      }
-      \vfill
-      \onslide<2->{
-        \item The OBMOP has some limitations: The owner is the only entity that enforces rules, and the abstraction doesn't guarantee things like freedom from deadlock.
-      }
-      \vfill
-      \onslide<3->{
-        \item This approach has potential: The authors were able to implement STM, CSP, agents, and actors in less than 500 total lines of code.
       }

### 2016-11-18 slides/slides.tex
Typeset atomic variable figures
Typeset remaining figures and started fixing conclusion

@@
 \usepackage[scale=2]{ccicons}
+\usepackage{color}
 \usepackage{graphicx}
@@
 \usepackage{multicol}
+\usepackage{pifont}
 \usepackage{tikz}
+\usetikzlibrary{positioning}
 
@@
   \maketitle
+
+  \section*{Introduction}
 
@@
         \centering
-        \includegraphics[width=160pt]{../figures/nonatomic}
+        \input{../figures/nonatomic.tikz}
       \end{column}
@@
         \centering
-        \includegraphics[width=160pt]{../figures/atomic}
+        \input{../figures/atomic.tikz}
       \end{column}
@@
     \centering
-
-    \includegraphics[width=220pt]{../figures/lock}
+    \input{../figures/lock.tikz}
   \end{frame}
@@
     \centering
-
-    \includegraphics[width=220pt]{../figures/multiple-locks}
+    \input{../figures/multiple-locks.tikz}
   \end{frame}
@@
     \centering
-
-    \includegraphics[width=220pt]{../figures/stm}
+    \input{../figures/stm.tikz}
   \end{frame}
@@
     \centering
-
-    \includegraphics[width=220pt]{../figures/agent}
+    \input{../figures/agent.tikz}
     \vfill
-    \textbf{Agents:} An isolated thread handles all operations on an object.~\footcite{Swalens2014}
+    \textbf{Agents:} An isolated thread wraps an object.~\footcite{Swalens2014}
   \end{frame}
@@
     \centering
-
-    \includegraphics[width=220pt]{../figures/csp}
+    \input{../figures/csp.tikz}
     \vfill
@@
     \centering
-
-    \includegraphics[width=220pt]{../figures/actor}
+    \input{../figures/actor.tikz}
     \vfill
@@
     \centering
+    \small
 
-    \includegraphics[width=360pt]{../figures/clojure-survey}
+    \newcommand{\n}{\textcolor{red}{\ding{55}}}
+    \newcommand{\y}{\textcolor{olive}{\ding{51}}}
+
+    \textbf{Pairwise combinations of four Clojure concurrency models:}
+    \vfill
+    \begin{tabular}{c | c c c c | c c c c}
+      & \multicolumn{4}{c|}{\textbf{Safety}} & \multicolumn{4}{|c}{\textbf{Liveness}} \\
+      & atoms & refs & agents & channels & atoms & refs & agents & channels \\
+      \hline
+      atoms & \n & \n & \n & \n & \y & \y & \y & \n \\
+      refs & \n & \y & \y & \n & \y & \y & \y & \n \\
+      agents & \y & \y & \y & \y & \y & \y & \y & \n \\
+      channels & \y & \y & \y & \y & \y & \y & \n & \n
+    \end{tabular}
   \end{frame}
@@
 
-  \subsection{Conclusions}
+  \subsection{Future work}
 
   \begin{frame}
-    \frametitle{Conclusions}
+    \frametitle{Future work}
 
@@
       \onslide<1->{
-        \item The language may ensure that some bad interactions don't happen.
+        \item Swalens et al.\ planned to identify composable ``building blocks'' (thread creation, message passing, etc.) that could be used to build common concurrency models.~\footcite{Swalens2014}
       }
@@
       \onslide<2->{
-        \item Sometimes, it's best to just use a different concurrency model.
+        \item Marr and D'Hondt proposed a unifying abstraction for high-level language virtual machines.~\footcite{Marr2012}
       }
@@
       \onslide<3->{
-        \item Future work would identify common ``building blocks'' (thread creation, message passing, etc.) that could be composed to express existing concurrency models and combinations of models.
+        \item Ziv et al.\ formalized a theory for safely composing concurrency control.~\footcite{Ziv2015}
       }

### 2016-11-18 slides/slides.tex
Added slide about threading and prepped some other stuff for the last practice talk

@@
 \usetikzlibrary{positioning}
+\usetikzlibrary{snakes}
 
@@
   \subsection{Concurrency}
+
+  \begin{frame}
+    \frametitle{Threads and processes}
+
+    \begin{figure}
+      \centering
+      \input{../figures/threads.tikz}
+    \end{figure}
+
+    \vfill
+
+    \begin{itemize}
+      \item \textbf{Threads} are independent sequences of operations.
+      \item \textbf{Processes} are instances of programs made up of one or more threads.
+    \end{itemize}
+  \end{frame}
 
@@
     \frametitle{Software transactional memory}
-    \centering
-    \input{../figures/stm.tikz}
-  \end{frame}
-
-  \begin{frame}
-    \frametitle{Software transactional memory}
 
@@
     \end{enumerate}
+  \end{frame}
+
+  \begin{frame}
+    \frametitle{Software transactional memory}
+    \centering
+    \input{../figures/stm.tikz}
   \end{frame}
@@
 
+  \subsection{Possible conflicts}
+
   \begin{frame}
-    \frametitle{Composability criteria}
+    \frametitle{Possible conflicts}
     \centering
@@
 
-    \textbf{Pairwise combinations of four Clojure concurrency models:}
     \vfill
@@
     \end{tabular}
-  \end{frame}
-
-  \subsection{Possible conflicts}
-
-  \begin{frame}
-    \frametitle{Possible conflicts}
-
+    \vfill
     \begin{itemize}
-      \onslide<1->{
-        \item A model reexecutes code containing another model that performs an irrevocable action.
+      \onslide<2->{
+        \item A model reexecutes code that performs an irrevocable action.
       }
-      \vfill
-      \onslide<2->{
+      \onslide<3->{
         \item A model reexecutes code that causes the reexecution to continually happen.
       }
-      \vfill
-      \onslide<3->{
-        \item A model that supports blocking operations is used within a model that does not support blocking operations.
+      \onslide<4->{
+        \item A model that supports blocking operations is used within a model that doesn't.
       }
-      \vfill
-      \onslide<4->{
-        \item A model does not guarantee safety or liveness by design, making composability impossible.
+      \onslide<5->{
+        \item A model does not guarantee safety or liveness by design.
       }
@@
 
-  \subsection{Future work}
+  \subsection{Ongoing work}
 
   \begin{frame}
-    \frametitle{Future work}
+    \frametitle{Ongoing work}
 
@@
       \onslide<1->{
-        \item Swalens et al.\ planned to identify composable ``building blocks'' (thread creation, message passing, etc.) that could be used to build common concurrency models.~\footcite{Swalens2014}
+        \item Composable ``building blocks'' (thread creation, message passing, etc.) that could be used to build common concurrency models~\footcite{Swalens2014}
       }
@@
       \onslide<2->{
-        \item Marr and D'Hondt proposed a unifying abstraction for high-level language virtual machines.~\footcite{Marr2012}
+        \item Unifying abstractions for high-level language virtual machines~\footcite{Marr2012}
       }
@@
       \onslide<3->{
-        \item Ziv et al.\ formalized a theory for safely composing concurrency control.~\footcite{Ziv2015}
+        \item Formal theories for safely composing concurrency control~\footcite{Ziv2015}
       }

### 2016-11-19 slides/slides.tex
Added numbering to table of contents
Made some last-minute changes

@@
 \usepackage{color}
+\usepackage{colortbl}
+\usepackage{diagbox}
 \usepackage{graphicx}
@@
 
+\defbeamertemplate{section in toc}{numbered titles}{%
+  \leavevmode
+  \parbox[t]{1.5em}{\inserttocsectionnumber.}%
+  \parbox[t]{\dimexpr\textwidth-1em\relax}{\inserttocsection}\par
+}
+
+\defbeamertemplate{subsection in toc}{indented titles}{%
+  \leavevmode
+  \parbox[t]{3em}{\hfill}%
+  \parbox[t]{\dimexpr\textwidth-1em\relax}{\inserttocsubsection}\par
+}
+
 % this actually isn't totally hideous:
@@
 \setbeamertemplate{frametitle continuation}{}
+\setbeamertemplate{section in toc}[numbered titles]
+\setbeamertemplate{subsection in toc}[indented titles]
 
@@
     \end{enumerate}
-
-    If $A \rightarrow B$ and $B \rightarrow C$, then $A \rightarrow C$.
 
@@
     \frametitle{Possible conflicts}
-    \centering
     \small
@@
     \begin{tabular}{c | c c c c | c c c c}
-      & \multicolumn{4}{c|}{\textbf{Safety}} & \multicolumn{4}{|c}{\textbf{Liveness}} \\
-      & atoms & refs & agents & channels & atoms & refs & agents & channels \\
+      & \multicolumn{4}{c|}{\textbf{Safety}} & \multicolumn{4}{|c}{\textbf{Liveness}} \tabularnewline
+      \diagbox[dir=NW,width=7em]{within}{using} & atoms & refs & agents & channels & atoms & refs & agents & channels \tabularnewline
       \hline
-      atoms & \n & \n & \n & \n & \y & \y & \y & \n \\
-      refs & \n & \y & \y & \n & \y & \y & \y & \n \\
-      agents & \y & \y & \y & \y & \y & \y & \y & \n \\
-      channels & \y & \y & \y & \y & \y & \y & \n & \n
+      atoms & \n & \n & \n & \n & \y & \y & \y & \n \tabularnewline
+      refs & \n & \y & \y & \n & \y & \y & \y & \n \tabularnewline
+      agents & \y & \y & \y & \y & \y & \y & \y & \n \tabularnewline
+      channels & \y & \y & \y & \y & \y & \y & \n & \n \tabularnewline
     \end{tabular}

### 2016-11-23 paper/paper.tex
Cleaned up presentation intro/background and typeset diagrams
Made some changes for the final draft

--- 2016-10-27 paper/paper.tex
@@
 \usepackage[backend=bibtex,style=numeric-comp]{biblatex}
-\usepackage{graphicx}
-
-\graphicspath{{../figures/}}
+\usepackage{tikz}
+\usetikzlibrary{positioning}
 
@@
 Modern programming languages generally provide several concurrency models to serve different requirements, and programmers benefit from being able to use them in tandem.
-We discuss challenges surrounding concurrency control and examine situations in which conflicts between models can occur.
-Additionally, we describe attempts to implement concurrency models on top of lower-level concurrency abstractions.
+We discuss challenges surrounding concurrent programming and examine situations in which conflicts between models can occur.
+Additionally, we describe attempts to identify features of common concurrency models and develop lower-level abstractions capable of supporting a variety of models.
 
 \end{abstract}
-
-\keywords{concurrency, parallelism, concurrency abstraction}
 
@@
 A web browser, for instance, might at any point be rendering documents in multiple tabs, transferring files, and handling user interaction.
-On a lower level, the operating system might be running several other applications, juggling background processes, and responding to events~\cite{Swalens2014}.
+On a lower level, the operating system might be running several other applications, juggling background processes, and responding to events.
 If every long-running process blocked other processes from proceeding, the system would be effectively unusable.
 
-Concurrency models enable programmers to reason about and describe interactions between concurrent processes.
-The web browser is a good example of several different models: The user interface layer might rely on an event loop, the rendering process might operate in shared memory, and suggestions from browsing history or a search engine might require parallel collection operations to achieve acceptable performance~\cite{Marr2012}.
+Processes themselves are often composed of multiple concurrent threads of execution.
+A processor can only execute one thread at a time, so multitasking is accomplished by rapidly switching between threads~\cite{Liu1973}.
+Although concurrent threads may appear to be executed simultaneously, truly parallel execution can only take place across multiple processors.
 
-Given that an application is likely to make use of more than one concurrency model, programmers would prefer that different types of models could be arbitrarily combined.
-From an engineering standpoint, composable models are also desir\-able---a virtual machine for one or more high-level languages should be able to support a variety of models without resorting to rough low-level approximations.
-However, different concurrency models are not necessarily composable and unexpected issues may arise when they interact.
-Recent work has attempted to identify unifying concurrency abstractions that would guarantee composability of different models and also provide underlying virtual machine support~\cite{Marr2009, Marr2012, Swalens2014, Ziv2015}.
+Concurrency models enable programmers to reason about concurrent tasks instead of low-level thread management.
+The web browser is a good example of how different models might be chosen to represent specific types of tasks: The user interface layer might rely on an event loop, the rendering process might operate in shared memory, and suggestions from browsing history or a search engine might require parallel collection operations to achieve acceptable performance~\cite{Marr2012}.
+
+Given that an application is likely to make use of more than one concurrency model, programmers would prefer that different types of models could safely interact.
+However, different models do not necessarily work well together, nor are they designed to.
+Recent work has attempted to identify common ``building blocks'' that could be used to compose a variety of models, possibly eliminating subtle problems when different models interact and allowing models to be represented at lower levels without resorting to rough approximations~\cite{Marr2009, Marr2012, Swalens2014, Ziv2015}.
 
@@
 
-Modern operating systems are expected to run many processes concurrently, and processes themselves are often composed of multiple concurrent threads of execution.
-A processor can only execute one thread at a time, so multitasking is accomplished by rapidly switching between threads~\cite{Liu1973}.
-Although concurrent threads may appear to be executed simultaneously, truly parallel execution can only take place across multiple processors.
+In a concurrent program, the history of operations may not be the same for every execution.
+An entirely sequential program could be proved to be correct by showing that its history (that is, the sequence in which its operations are performed) always yields a correct result.
+For a concurrent program to be proved correct, though, it must be clear that all possible histories always yield a correct result.
+Furthermore, concurrent programming frequently involves the use of shared resources that require coordinated access and manipulation.
 
-Concurrency models can abstract over implementation details, allowing programmers to reason in terms of asynchron\-ous tasks and independent components instead of low-level thread management.
-In addition to making a program more understandable, higher-level models afford a degree of flexibility: Regardless of whether a set of concurrent operations is executed in sequence on one thread, in multiple threads on the same processor, or in parallel on multiple processors, the result will be the same.
+Figure~\ref{figure:single} illustrates a simple program in which a single thread reads ($=$) and writes ($\leftarrow$) values on a variable $x$, denoted by a segmented bar.
+The program satisfies an intuitive model of how variables should behave---each time $x$ is read, the value returned is equal to the value most recently written.
+Because the thread has exclusive control of $x$, that assumption is valid.
 
-\subsection{Concurrency}
+\begin{figure}[ht]
+  \centering
+  \resizebox{\linewidth}{!}{\input{../figures/single-thread-read-write.tikz}}
+  \caption{A single thread reads from and writes to a variable.}
+\label{figure:single}
+\end{figure}
 
-For operations to be executed concurrently, they must be able to be executed out of order or in partial order.
-Lamport, in his foundational work on distributed systems~\cite{Lamport1978}, formalized this by defining a ``happens before'' relation (denoted by ``$\rightarrow$'' and its negation by ``$\nrightarrow$'') on a set of operations: If $A$ and $B$ are operations in the same process and $A$ occurs before $B$, or if $A$ is the sending of a message by one process and $B$ is the receipt of the same message by another process, then $A \rightarrow B$.
-Two operations $A$ and $B$ are said to be concurrent if $A \nrightarrow B$ and $B \nrightarrow A$.
+In Figure~\ref{figure:multiple}, on the other hand, two threads concurrently read and write on $x$.
+Because the operations of the threads are interleaved, a read operation on $x$ may not yield a value that matches the value of that thread's most recent write.
+The value may even be inconsistent between consecutive reads.
+If a single thread incorrectly assumed that it had exclusive control of $x$, this behavior could lead to inconsistent or incorrect results.
+Similar errors that arise due to an unintentional dependency on execution order are commonly referred to as race conditions.
 
-While the ``happens before'' relation can be used to determine whether operations can be executed concurrently, it does not guarantee the correctness of the results.
-In a non-concurrent (entirely sequential) program, it would be sufficient to show that the program is correct after the execution of each operation and therefore correct upon termination.
-In other words, the program could be proved to be correct by proving that its history (that is, the sequence in which its operations are performed) yields a correct result.
+\begin{figure}[ht]
+  \centering
+  \resizebox{\linewidth}{!}{\input{../figures/multiple-threads-read-write.tikz}}
+  \caption{Two threads concurrently read from and write to a variable.}
+\label{figure:multiple}
+\end{figure}
+
 Consistency models guarantee the correctness of a program by defining a set of all allowed histories, or histories that lead to a correct result~\cite{Ziv2015}.
@@
 
-\begin{figure}[h]
-  \centering
-  \includegraphics{single-thread-read-write}
-  \caption{A single thread writes to and reads from a variable.}
-\label{figure:single}
-\end{figure}
-
-Figure~\ref{figure:single} illustrates a simple program in which a single thread reads and writes numbers from a variable $x$.
-The value of $x$ over time is represented by a segmented bar.
-Write operations are denoted by an arrow pointing toward the bar and read operations by an arrow pointing away.
-The program satisfies an intuitive model of how variables should behave---each time $x$ is read, the value is equal to the value of the most recent write.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics{multiple-threads-read-write}
-  \caption{Two threads concurrently write to and read from a variable.}
-\label{figure:multiple}
-\end{figure}
-
-In a concurrent program, the history of operations may not be the same for every execution.
-As a result, assumptions about correctness that are true for a sequential program may be violated in a concurrent program.
-This is illustrated in Figure~\ref{figure:multiple}, in which two threads concurrently write from and read to a variable $x$ concurrently.
-Because the operations of the threads are interleaved, a read operation on $x$ may not yield a value that matches the value of that thread's most recent write.
-The value may even be inconsistent between consecutive reads.
-If a single thread were to incorrectly assume that it had exclusive control of $x$, this behavior would appear incorrect and could lead to race conditions, errors that arise due to unintentional dependency on execution order.
-
-\subsection{Linearizability}
+\subsection{Linearizable consistency}
 
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
   \centering
-  \includegraphics{noninstantaneity}
+  \resizebox{0.63\linewidth}{!}{\input{../figures/noninstantaneity.tikz}}
   \caption{Two threads concurrently (and noninstantaneously) write to and read from a variable.}
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
   \centering
-  \includegraphics{linearizability}
+  \resizebox{0.63\linewidth}{!}{\input{../figures/linearizability.tikz}}
   \caption{A single thread performs an atomic write operation on a variable.}
@@
 
-\subsection{Serializability}
+\subsection{Sequential and causal consistency}
+
+For operations to be executed concurrently, they must be able to be executed out of order or in partial order.
+Lamport, in his foundational work on distributed systems~\cite{Lamport1978}, formalized this by defining a ``happens before'' relation ($\rightarrow$) and its negation ``does not happen before'' ($\nrightarrow$) on a set of operations: If $A$ and $B$ are operations in the same thread and $A$ occurs before $B$, or if $A$ is the sending of a message by one thread and $B$ is the receipt of the same message by another thread, then $A \rightarrow B$.
+If $A \rightarrow B$ and $B \rightarrow C$, then $A \rightarrow C$.
+Two operations $A$ and $B$ are said to be concurrent if $A \nrightarrow B$ and $B \nrightarrow A$.
+
+The first condition of the ``happens before'' relation, that $A$ and $B$ are ordered within the same thread, describes sequential consistency.
+Sequential consistency simply requires that operations in a thread take place in that thread's order.
+Causal consistency, described by the second condition of the ``happens before'' relation, enforces the intuitive requirement that an effect cannot take place before its cause.
+
+\subsection{Serializable consistency}
 
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
   \centering
-  \includegraphics{serializability}
+  \resizebox{0.63\linewidth}{!}{\input{../figures/serializability.tikz}}
   \caption{Two threads perform a set of serializable read and write operations on a variable.}
@@
 Figure~\ref{figure:serializability} illustrates a set of serializable read and write operations.
-Even though the exeuctions of each operation occur out of order and even intersect, the events themselves can be arranged serially: $x \leftarrow 1$, $x = 1$, $x \leftarrow 2$, $x = 2$.
+Even though the executions of each operation occur out of order and even intersect, the events themselves can be arranged serially.
 
@@
 Because it does not place bounds on time or order, events may happen out of sequence, and, unlike linearizability, serializability is not a composable guarantee.
-Serializability is useful, though, as a guarantee of isolation~\cite{Haerder1983}: While a serializable set of operations is being executed, it appears to be the only set of operations being executed.
+Serializability is useful, though, as a guarantee of isolation: While a serializable set of operations is being executed, it appears to be the only set of operations being executed.
 For histories to be serializable, then, they must not overlap and cannot interfere with each other.
 
-Together, linearizability and serializability guarantee strict serializability, in which a history is equivalent to a serial execution order and that serial order corresponds to the execution order in real time.
+Linearizability and serializability together guarantee strict serializability, in which a history is equivalent to a serial execution order and that serial order corresponds to the execution order in real time.
 Given that, linearizability can actually be described as strict serializability restricted to single operations on single objects~\cite{Herlihy1990}.
@@
 
-Modern programming languages generally provide several different concurrency models, each with their own terminology and specific implementation details.
-It would be impractical to discuss them all, but the most commonly used models can be adequately described by generalization~\cite{Swalens2014}.
+Modern programming languages generally provide several different concurrency models as tools to reason about concurrent tasks.
+Swalens et.\ al, in a survey of concurrency models in the Clojure programming language, group commonly used concurrency models into four broad categories: atomic variables, software transactional memory, communicating threads, and proxies~\cite{Swalens2014}.
 
@@
 Atomic operations are often implemented in hardware and guarantee linearizability at the lowest level possible.
-\code{fetch-and-add}, another atomic operation, is a common processor instruction that executes multiple hardware-level operations: The value old value is copied from a location in memory into a temporary register, addition is performed on the value in the temorary register, and the new value is stored at the original location.
+\code{fetch-and-add}, another atomic operation, is a common processor instruction that executes multiple hardware-level operations: The old value is copied from a location in memory into a temporary register, addition is performed on the value in the temorary register, and the new value is stored at the original location.
 
@@
 
-Many higher-level languages (such as Clojure~\cite{Swalens2014}) provide atomic references that can be read by an atomic dereferencing operation and written by an atomic swap operation.
-Such operations may provide convenience features such as automatic retries.
+While lock-based programming is effective, dealing with multiple locks and complex interactions can be cumbersome.
+One common problem in lock-based programming is resource contention, in which multiple threads vie for control of multiple resources.
+Suppose, for example, that two threads $A$ and $B$ both require resources $x$ and $y$.
+$A$ might acquire $x$ and $B$ might acquire $y$.
+Both would then halt, waiting to acquire a lock on the second needed resource.
+This type of scenario is commonly referred to as deadlock.
 
@@
 
-Software transactional memory (STM) presents an alternative to semaphores and locks by allowing multiple concurrent operations to write to a shared location in memory.
+Software transactional memory (STM) presents an alternative to lock-based programming by allowing multiple concurrent operations to write to a shared location in memory without securing any kind of lock.
 STM is an example of optimistic concurrency control: Each thread executes a series of read and write operations (called a transaction) on the shared memory without considering the activity of other threads.
@@
 Only when a transaction completes successfully will changes will become visible to other threads.
+Wrapping a set of operations within a transaction eliminates the need to secure locks, leading to more simple programs.
 Practically, STM is useful in any situation in which a shared object needs to be accessed and modified by multiple threads~\cite{Swalens2014}.
@@
 STM also offers significant advantages over lock-based programming in terms of composability.
-Given that executing two operations within a transaction yields a larger atomic operation, Harris et al.\ describe a STM that allows any combination of atomic operations to be composed into a larger atomic operation~\cite{Harris2005}.
+Because a transaction can be viewed as a single linearizable operation, STM can allow any combination of linearizable operations to be composed into a larger linearizable operation.
+Transactions also enforce modularity by hiding implementation details of the operations within~\cite{Harris2005}.
 
@@
 Specifically, a transaction cannot execute any operation that cannot be undone or retried, such as writing to disk or performing a network request.
-STM can also result in a decrease in performance caused by the overhead of maintaining the log and aborting and retrying transactions.
+Operations that block indefinitely may also cause a transaction to fail continuously.
+STM can result in a decrease in performance caused by the overhead of maintaining the log and aborting and retrying transactions.
 
@@
 
-Other concurrently models avoid the use of shared memory (and the problems that arise from doing so) by restricting threads to private memory.
-Threads communicate strictly by message passing, which avoids race conditions.
-A variety of different communicating thread models exist and have long been a solution to describing concurrent operations.
-Practical use cases for communicating threads include communication with external systems or running an event loop~\cite{Swalens2014}.
+Other concurrently models avoid the use of shared memory by restricting threads to private memory.
+Threads communicate strictly by message passing, which avoids issues such as race conditions.
+Sending messages and reacting to received messages are inherently asynchronous processes, so communicating threads have long been a solution to describing concurrent operations.
+Limiting thread communication to message passing also enforces modularity and limits possible negative side effects.
+A wide variety of communicating thread models exist and often run event loops or handle communication with external systems~\cite{Swalens2014}.
 
-Communicating sequential processes (CSP), one of the oldest communicating thread models, describes systems in terms of independent processes that communicate through predefined channels~\cite{Hoare1978}.
+Communicating sequential processes (CSP), one of the oldest communicating thread models, began as a method of formally describing actions in concurrent systems.
+CSP describes systems in terms of independent processes that communicate through predefined channels~\cite{Hoare1978}.
 CSP is notable in that messages are passed synchronously; that is, a sending process will block until a complementary operation is executed on the receiving process~\cite{Swalens2014}.
+CSP is still widely used in the specification, implementation, and testing of safety-critical systems.
 
-Other models, most notably the actor model, rely on asynchronous message passing to specific entities~\cite{Agha1986}.
+Other models, most notably the actor model, rely on asynchronous message passing to specific entities (actors) rather than named channels~\cite{Agha1986}.
 Actors can only communicate via asynchronous message passing.
@@
 
-Futures and promises are two of the most common proxy models, though the terms are frequently used interchangeably (along with ``delayed,'' ``deferred,'' and ``eventual'').
-Generally, futures are resolved to the result of the completed operation.
+Futures and promises are two of the most common proxy models.
+The terms are frequently used interchangeably, along with ``delayed,'' ``deferred,'' and ``eventual.'' Generally, futures are resolved to the result of the completed operation.
 The result is then accessed implicitly; any use of the future will return its value.
-Promises, on the other hand, are created as independent objects and require the result to be accessed explicitly.
+Promises differ in that they are created as independent objects and require the result to be accessed explicitly.
 
 In practice, proxies are used to execute long-running operations such as rendering or network requests~\cite{Swalens2014} without blocking other operations.
-Promise objects, which can be easily passed and composed, generally offer more flexibility than constructs such as asynchronous callbacks~\cite{Kambona2013}.
+Promise objects, which can be easily passed and composed, generally offer more flexibility than constructs such as asynchronous callbacks.
 
-\section{Proposed abstractions}
+\section{Composability challenges}
 
-\subsection{Asynchronous events}
+While concurrency models guarantee their own correctness, issues may arise when different models are used together.
+For example, an implementation of an STM that guarantees linearizability assumes that all shared resources are managed by the STM~\cite{Shavit1995}.
+If such a resource is modified outside of the STM, an unexpected interleaving of operations could result.
 
-Ziarek et al.~\cite{Ziarek2011} introduce asynchronous events as a possible abstraction. Their examples are heavily tied to Concurrent ML and might be difficult to integrate.
+Swalens et al.~\cite{Swalens2014} attempted to identify issues that arise when different concurrency models are combined, hoping that future work could identify common ``building blocks'' that could be used to compose a variety of concurrency models.
+The study surveyed all of the concurrency models available in the Clojure programming language and used them in tandem with one another to see what types of correctness issues were encountered.
+
+\subsection{Correctness criteria}
+
+In the study, safety and liveness were used as the two criteria for evaluating the correctness of a combination of models.
+Together, the two have historically been used to prove the correctness of distributed systems~\cite{Lamport1977}.
+
+Informally, safety guarantees that ``nothing bad will happen.'' In other words, given a correct input, a program will not produce an incorrect result~\cite{Swalens2014}.
+This property is also referred to as partial correctness.
+Safety is generally achieved in the management of shared resources.
+STM, for instance, only allows shared memory to be accessed through transactions; communicating threads only allow data to be shared through message passing.
+
+Liveness guarantees that ``something good will eventaully happen,'' or that a program will eventually terminate if its input is correct.
+Taken together, safety and liveness are referred to as total correctness---given a correct input, a program will terminate with the correct output~\cite{Swalens2014}.
+Deadlocks and livelocks are the primary obstacle to liveness and can arise when models are combined.
+
+\subsection{Possible conflicts}
+
+To discover the types of conflicts that might arise when combining different models, the study examined all pairwise combinations of Clojure's models.
+Using every model within every other model uncovered the following types of conflicts:
+
+\begin{itemize}
+  \item A model might reexecute code containing another concurrency model that performs an irrevocable action.
+For instance, if a transaction were to contain an operation that sent a message on a channel, reexecutions of the transaction would cause the message to be sent repeatedly.
+
+  \item A model might reexecute code that causes the reexecution to continually happen.
+The authors of the study observed that multiple STM transactions might consistently conflict with another and never succeed.
+
+  \item A model that supports blocking operations might be used within a model that does not expect blocking operations.
+The authors of the study provided mutually recursive futures as an example---if two futures each attempted to read the value of the other, a deadlock would occur.
+
+  \item A model might not guarantee safety or liveness by design, making composability impossible.
+Perhaps obviously, a model that does not guarantee safety or liveness would violate the safety guarantee of an outer model unless such a conflict was explicitly prevented.
+\end{itemize}
+
+The study noted that some types of bad interactions were already prevented by Clojure; for instance, sending a message to an agent from within a transaction is delayed until the transaction succeeds.
+The authors noted that similar safeguards could prevent other negative interactions.
+However, preventing every type of negative interaction between models might undermine the purpose of the models to begin with; channels, as mentioned previously, block by design and are inherently incompatible with other models.
+Therefore, further research would likely seek to decompose concurrency models into common elements and provide a method to compose those elements safely and efficiently.
+
+\section{Unifying abstractions}
+
+Other research into composable concurrency models has focused on the development of underlying abstractions capable of supporting a wide variety of models.
+Instead of programmers using composable ``building blocks'' to construct models, these abstractions would make use of those common elements at a lower level to improve performance and interoperability.
 
@@
 
-Marr and D'Hondt~\cite{Marr2012} surveyed a variety of concurrent and parallel programming concepts to identify a unifying concurrency abstraction that could be implemented at the virtual machine level.\footnote{A 2009 survey by Marr, Haupt, and D'Hondt~\cite{Marr2009} found that virtual machines at the time did not provide intrinsic support for more than one specific model.} A unifying abstraction would enable each concurrency model in a high-level language to be represented in machine code without sacrificing performance or semantics.
-To identify concepts relevant for a multi-language virtual machine, Marr and D'Hondt selected four questions:
+Marr and D'Hondt~\cite{Marr2012} surveyed a variety of concurrent and parallel programming concepts such as immutability, critical sections, ownership, and common concurrency models to identify a unifying concurrency abstraction that could be implemented in a high-level language virtual machine.\footnote{A 2009 survey by Marr, Haupt, and D'Hondt~\cite{Marr2009} found that virtual machines at the time did not provide intrinsic support for more than one specific model.} A unifying abstraction would enable each concurrency model in a high-level language to be represented in machine code without sacrificing performance or semantics.
+For instance, a concurrency model in Clojure could be accurately represented by an underlying abstraction in the Java Virtual Machine, and code written in the Java language could interact with that model correctly.
+To identify the concepts that would need to be supported at the virtual machine level, Marr and D'Hondt selected four questions:
 
@@
   \item Can the concept be reasonably implemented as a library?
-(A library implementation may suffer from loss of performance or semantics to an extent that the concept would warrant virtual machine inclusion.)
+A library implementation may suffer from loss of performance or semantics to an extent that the concept would warrant virtual machine inclusion.
 
   \item Does the concept require runtime support to guarantee its semantics?
-(Semantic guarantees may be enforced by a compiler but not the virtual machine.
-For instance, immutable objects in one language may not be immutable in machine code, and therefore may be mutated by another language.)
+Semantic guarantees may be enforced by a compiler but not the virtual machine.
+For instance, immutable objects in one language may not be immutable in machine code, and therefore may be mutated by another language.
 
   \item Would runtime support benefit performance?
-(Deeper knowledge of the concept may enable the virtual machine to better optimize.
-Some concepts, though, may require knowledge of the underlying hardware to realize any performance improvements; such concepts would not necessarily benefit from runtime support.)
+Deeper knowledge of the concept may enable the virtual machine to better optimize.
+Some concepts, though, may require knowledge of the underlying hardware to realize any performance improvements; such concepts would not necessarily benefit from runtime support.
 
@@
 \begin{itemize}
-  \item Managed mutation: Concurrency models impose restrictions on how state can be modified; therefore, mutation must be able to be flexibly managed.
+  \item Managed mutation: A concurrency model may impose rules on how some state can be modified; therefore, mutation must be able to be flexibly managed.
 
@@
 
-  \item Ownership: Some owning entity usually regulates mutation and execution, so ownership should be supported in a way that allows for adaptable mutation and execution rules.
+  \item Ownership: An owning entity regulates mutation and execution, so ownership should be supported in a way that allows for adaptable mutation and execution rules.
 
   \item Leveled reflection: Reflection allows a program to examine its own structure and behavior at runtime.
-To guarantee safety, there needs to be a distinction between language-level reflection (restricted reflection operations supported by the high-level language) and meta-level reflection (unrestricted reflection operations on the low-level abstraction).
+To guarantee safety, there needs to be a distinction between language-level reflection (reflection operations supported by the high-level language) and meta-level reflection (reflection operations on the low-level abstraction).
 
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
   \centering
-  \includegraphics[width=224pt]{obmop}
+  \resizebox{0.63\linewidth}{!}{\input{../figures/obmop.tikz}}
   \caption{Marr and D'Hondt's object-based meta-object protocol~\cite{Marr2012}.}
@@
 
-In more words: There's a lot of work on the correctness of concurrent operations.
-With some care, concurrency models can be composed.
-There's some work on deeper abstractions, but it's definitely ongoing and more is likely to follow soon.
+There are a rich variety of concurrency models that can be used to represent concurrent tasks, abstract over low-level details, and ensure that concurrent applications perform correctly.
+Recent research has attempted to identify ways in which these models can be safely used together, common features of these models, and underlying abstractions capable of representing many different models.
+
+Future work will likely explore ways in which concurrency models can be decomposed and recomposed from common elements.
+This, in turn, would lead to the development of more generalized ways of describing concurrency models and perhaps additional unifying abstractions.
+As Marr and D'Hondt note, a ``silver bullet'' does not exist.
+However, there is certainly promise in searching for better ways to represent concurrent operations.
 
@@
 
-Thanks to Elena Machkasova and K.~K. Lamberty for their patience and feedback.
+Thanks to Elena Machkasova, K.~K. Lamberty, and Matthew Justin for their suggestions and feedback.
 

### 2016-11-30 paper/paper.tex
Did another round of revisions; unsuccessfully attempted to compile a list of things worse than BibTeX

@@
 However, different models do not necessarily work well together, nor are they designed to.
-Recent work has attempted to identify common ``building blocks'' that could be used to compose a variety of models, possibly eliminating subtle problems when different models interact and allowing models to be represented at lower levels without resorting to rough approximations~\cite{Marr2009, Marr2012, Swalens2014, Ziv2015}.
+Recent work has attempted to identify common ``building blocks'' that could be used to compose a variety of models, possibly eliminating subtle problems when different models interact and allowing models to be represented at lower levels without resorting to rough approximations~\cite{Marr2012, Swalens2014, Ziv2015}.
 
@@
 
-Figure~\ref{figure:single} illustrates a simple program in which a single thread reads ($=$) and writes ($\leftarrow$) values on a variable $x$, denoted by a segmented bar.
-The program satisfies an intuitive model of how variables should behave---each time $x$ is read, the value returned is equal to the value most recently written.
-Because the thread has exclusive control of $x$, that assumption is valid.
+Consistency models, restrictions on possible execution histories, can guarantee that a program will produce a correct result~\cite{Ziv2015}.
+If an execution of a program follows an allowed history, that execution is said to be consistent; if not, it is said to be inconsistent.
+If every possible execution of the program is guaranteed to follow a history allowed by a consistency model, that program conforms to that consistency model~\cite{Kingsbury2014}.
 
@@
   \centering
-  \resizebox{\linewidth}{!}{\input{../figures/single-thread-read-write.tikz}}
+  \resizebox{0.95\linewidth}{!}{\input{../figures/single-thread-read-write.tikz}}
   \caption{A single thread reads from and writes to a variable.}
@@
 
-In Figure~\ref{figure:multiple}, on the other hand, two threads concurrently read and write on $x$.
-Because the operations of the threads are interleaved, a read operation on $x$ may not yield a value that matches the value of that thread's most recent write.
-The value may even be inconsistent between consecutive reads.
-If a single thread incorrectly assumed that it had exclusive control of $x$, this behavior could lead to inconsistent or incorrect results.
-Similar errors that arise due to an unintentional dependency on execution order are commonly referred to as race conditions.
+Figure~\ref{figure:single} illustrates a simple program in which a single thread $A$ reads ($=$) and writes ($\leftarrow$) values on a variable $x$, denoted by a segmented bar containing the value of $x$ over time.
+The program satisfies an intuitive model of how variables should behave---each time $x$ is read, the value returned is equal to the value most recently written and the consistency model holds.
+If, however, a read on $x$ failed to return the most recently written value, the consistency model would be violated.
 
@@
   \centering
-  \resizebox{\linewidth}{!}{\input{../figures/multiple-threads-read-write.tikz}}
+  \resizebox{0.85\linewidth}{!}{\input{../figures/multiple-threads-read-write.tikz}}
   \caption{Two threads concurrently read from and write to a variable.}
@@
 
-Consistency models guarantee the correctness of a program by defining a set of all allowed histories, or histories that lead to a correct result~\cite{Ziv2015}.
-If an execution of a program follows an allowed history, the execution is consistent.
-If not, the execution is inconsistent.
-If every possible execution of the program follows an allowed history, the program conforms to the consistency model.
+In Figure~\ref{figure:multiple}, two threads $A$ and $B$ concurrently read and write on $x$.
+Because the operations of the threads are interleaved, a read operation on $x$ may not yield a value that matches the value of that thread's most recent write.
+The value may even be inconsistent between consecutive reads.
+If a single thread assumed that it had exclusive control of $x$, incorrectly applying the single variable consistency model, this behavior would appear to be inconsistent.
+Similar errors that arise due to an unintentional dependency on execution order are commonly referred to as race conditions.
 
@@
 Even writing to a location in cache memory, an operation measured in nanoseconds, is not truly instantaneous.
-That an operation may be completed at a time significantly later than the time at which it was invoked introduces uncertainty into a history of operations---the sequence of execution may be influenced by the time it takes for messages to travel.
+The fact that operations are completed at some time after they are invoked introduces uncertainty into a history of operations---the sequence of execution may be influenced by the time it takes for messages to travel~\cite{Kingsbury2014}.
 
@@
   \centering
-  \resizebox{0.63\linewidth}{!}{\input{../figures/noninstantaneity.tikz}}
+  \resizebox{0.8\linewidth}{!}{\input{../figures/noninstantaneity.tikz}}
   \caption{Two threads concurrently (and noninstantaneously) write to and read from a variable.}
@@
 
-Consider a third example (Figure~\ref{figure:time}) in which two threads interact with a variable $x$ noninstantanously.
+Consider a third example (Figure~\ref{figure:time}) in which two threads $A$ and $B$ interact with a variable $x$ noninstantanously.
 Invocation is denoted by the leftmost point of an event and completion by the rightmost point.
 The dotted line indicates the instant at which the operation takes effect.
-The lower thread invokes a read while the value of $x$ is set to $1$.
+The lower thread invokes a read while the value of $x$ is set to $2$.
 The upper thread invokes and completes a write between the time that the read is invoked and the value is actually read.
-The read operation then returns $2$ even though $x$ was equal to $1$ at the time it was invoked.
+The read operation then returns $3$ even though $x$ was equal to $2$ at the time it was invoked.
 
 While travel time may introduce ambiguity, there are still some restrictions on possible sequences of events.
-Specifically, an operation cannot take effect before its invocation, nor can it take effect after its completion.
-Therefore, assuming that all threads share some single global state and that operations on that state take place atomically, each operation will appear to take effect atomically at some point after it is invoked and before it is completed.
+Specifically, an operation cannot take effect before its invocation, nor can it take effect after its completion; rather, it will appear to take effect atomically at some point after it is invoked and before it is completed.
+
+This consistency model is referred to as linearizability (also referred to as atomicity or indivisibility), and it guarantees that the completion of a single operation on a single object will appear to the rest of the system to be instantaneous~\cite{Herlihy1990}.
+In other words, even though linearizable operations are executed concurrently and take time, they appear to happen in a simple linear order.
 
@@
   \centering
-  \resizebox{0.63\linewidth}{!}{\input{../figures/linearizability.tikz}}
+  \resizebox{0.76\linewidth}{!}{\input{../figures/linearizability.tikz}}
   \caption{A single thread performs an atomic write operation on a variable.}
@@
 
-This consistency model is referred to as linearizability (also referred to as atomicity or indivisibility), and it guarantees that the completion of an operation on a single object will appear to the rest of the system to be instantaneous~\cite{Herlihy1990}.
-In other words, even though linearizable operations are executed concurrently and take time, they appear to happen in a simple linear order.
 Figure~\ref{figure:linearizability} demonstrates an atomic write on a variable $x$: Although the operation takes time, the entire operation (denoted by the shaded area) can be collapsed into one apparently instantaneous event.
-
 A useful consequence of linearizability is that the results of an operation must be visible as soon as the operation is complete.
@@
   \centering
-  \resizebox{0.63\linewidth}{!}{\input{../figures/serializability.tikz}}
+  \resizebox{0.76\linewidth}{!}{\input{../figures/serializability.tikz}}
   \caption{Two threads perform a set of serializable read and write operations on a variable.}
@@
 
-Serializability alone is a fairly weak consistency model.
-Because it does not place bounds on time or order, events may happen out of sequence, and, unlike linearizability, serializability is not a composable guarantee.
+Serializability alone is a fairly weak consistency model because it does not place any restrictions on time or order~\cite{Kingsbury2014}.
+Events may happen out of sequence, and, unlike linearizability, serializability is not a composable guarantee.
 Serializability is useful, though, as a guarantee of isolation: While a serializable set of operations is being executed, it appears to be the only set of operations being executed.
@@
 Atomic variables can only be read and mutated by operations that are guaranteed to be linearizable.
-For example, the atomic operation \code{compare-and-swap} compares the current value of a variable to a given value and only performs a write if the values match~\cite{Swalens2014}.
-\code{compare-and-swap} ensures that a new value cannot be written based on outdated information.
-Suppose that a thread $A$ reads a variable $x$ and begins computing a new value.
+For example, the atomic operation \code{compare-and-swap} compares the current value of a variable to a given value and only performs a write if the values are equal~\cite{Swalens2014}.
+\code{compare-and-swap} ensures that a new value cannot be written based on outdated information: Suppose that a thread $A$ reads a variable $x$ and begins computing a new value.
 At the same time, another thread $B$ modifies $x$.
@@
 Atomic variables often serve as the basis for higher-level concurrency control mechanisms such as semaphores and locks, which in turn support the implementation of concurrent data structures.
-A counting semaphore is an atomic variable that serves as a record of how many units of a shared resource are available.
+A counting semaphore is an atomic variable that serves as a record of how many units of a shared resource, such as a connection or a variable, are available.
 By taking advantage of atomic operations, the semaphore can be safely updated and will reliably determine whether a shared resource can be used.
 A single thread can then test if a resource is available and acquire it before proceeding, thereby preventing race conditions~\cite{Swalens2014}.
-A lock (also referred to as a mutex) guarantees mutual exclusion, which requires that two threads cannot operate on a shared object at the same time.
+A lock, also referred to as a mutex, guarantees mutual exclusion, which requires that two threads cannot operate on a shared object at the same time.
 Locks are commonly implemented by a binary semaphore (that is, a semaphore that simply indicates whether a single resource is available).
@@
 However, the requirement that transactions must be able to be aborted and retried places constraints on the set of allowed operations.
-Specifically, a transaction cannot execute any operation that cannot be undone or retried, such as writing to disk or performing a network request.
-Operations that block indefinitely may also cause a transaction to fail continuously.
-STM can result in a decrease in performance caused by the overhead of maintaining the log and aborting and retrying transactions.
+Specifically, a transaction cannot execute any operation that cannot be undone, such as writing to disk or performing a network request.
+Operations that block indefinitely, like waiting to acquire a lock, may also cause a transaction to fail continuously.
+STM can also result in a decrease in performance caused by the overhead of maintaining the log and aborting and retrying transactions.
 
@@
 Limiting thread communication to message passing also enforces modularity and limits possible negative side effects.
-A wide variety of communicating thread models exist and often run event loops or handle communication with external systems~\cite{Swalens2014}.
+A wide variety of communicating thread models exist and are often used to implement event loops or handle communication with external systems~\cite{Swalens2014}.
 
@@
 Other models, most notably the actor model, rely on asynchronous message passing to specific entities (actors) rather than named channels~\cite{Agha1986}.
-Actors can only communicate via asynchronous message passing.
-Upon receiving a message, an actor can choose to send messages to other actors, create new actors, and determine the behavior used for the next received message.
+Actors can only communicate by asynchronous message passing.
+Upon receiving a message, an actor can choose to perform an operation on a resource that it controls, send messages to other actors, create new actors, determine the behavior used for the next received message, or ignore the message entirely.
 
@@
 In practice, proxies are used to execute long-running operations such as rendering or network requests~\cite{Swalens2014} without blocking other operations.
-Promise objects, which can be easily passed and composed, generally offer more flexibility than constructs such as asynchronous callbacks.
+Promise objects, which can be easily passed and composed, offer more flexibility than constructs such as asynchronous callbacks.
 
@@
 
-While concurrency models guarantee their own correctness, issues may arise when different models are used together.
+Even if concurrency models guarantee their own correctness, issues may arise when different models are used together.
 For example, an implementation of an STM that guarantees linearizability assumes that all shared resources are managed by the STM~\cite{Shavit1995}.
@@
 Swalens et al.~\cite{Swalens2014} attempted to identify issues that arise when different concurrency models are combined, hoping that future work could identify common ``building blocks'' that could be used to compose a variety of concurrency models.
-The study surveyed all of the concurrency models available in the Clojure programming language and used them in tandem with one another to see what types of correctness issues were encountered.
+The study surveyed all of the concurrency models available in the Clojure programming language and used them in within one another to see what types of correctness issues were encountered.
 
@@
 Taken together, safety and liveness are referred to as total correctness---given a correct input, a program will terminate with the correct output~\cite{Swalens2014}.
-Deadlocks and livelocks are the primary obstacle to liveness and can arise when models are combined.
+Deadlocks (in which execution is blocked) and livelocks (in which execution continues indefinitely but never makes progress) are the primary obstacle to liveness and can arise when models are combined.
 
@@
 To discover the types of conflicts that might arise when combining different models, the study examined all pairwise combinations of Clojure's models.
-Using every model within every other model uncovered the following types of conflicts:
+Using every model within every other model (for example, manipulating an atomic variable within an STM transaction) uncovered the following types of conflicts:
 
@@
   \item A model might reexecute code that causes the reexecution to continually happen.
-The authors of the study observed that multiple STM transactions might consistently conflict with another and never succeed.
+The authors of the study observed that, especially when many transactions are being executed at once, a large STM transaction might consistently conflict with another and never succeed.
 
@@
 
-The study noted that some types of bad interactions were already prevented by Clojure; for instance, sending a message to an agent from within a transaction is delayed until the transaction succeeds.
-The authors noted that similar safeguards could prevent other negative interactions.
+The study noted that some types of bad interactions are prevented by Clojure; for instance, sending a message to an agent from within a transaction is delayed until the transaction succeeds.
+The study also noted that similar safeguards could prevent other negative interactions.
 However, preventing every type of negative interaction between models might undermine the purpose of the models to begin with; channels, as mentioned previously, block by design and are inherently incompatible with other models.
-Therefore, further research would likely seek to decompose concurrency models into common elements and provide a method to compose those elements safely and efficiently.
+With that in mind, further research would likely seek to decompose concurrency models into common elements and provide a method to compose those elements safely and efficiently.
 
@@
 
-Marr and D'Hondt~\cite{Marr2012} surveyed a variety of concurrent and parallel programming concepts such as immutability, critical sections, ownership, and common concurrency models to identify a unifying concurrency abstraction that could be implemented in a high-level language virtual machine.\footnote{A 2009 survey by Marr, Haupt, and D'Hondt~\cite{Marr2009} found that virtual machines at the time did not provide intrinsic support for more than one specific model.} A unifying abstraction would enable each concurrency model in a high-level language to be represented in machine code without sacrificing performance or semantics.
+Marr and D'Hondt~\cite{Marr2012} surveyed a variety of concurrent and parallel programming concepts such as immutability, critical sections, ownership, and common concurrency models to identify a unifying concurrency abstraction that could be implemented in a high-level language virtual machine.
+A unifying abstraction would enable each concurrency model in a high-level language to be represented in machine code without sacrificing performance or semantics.
 For instance, a concurrency model in Clojure could be accurately represented by an underlying abstraction in the Java Virtual Machine, and code written in the Java language could interact with that model correctly.
@@
 
-  \item Is the concept already supported by a common virtual machine like the Java Virtual Machine (JVM) or the Common Language Runtime (CLR)?
+  \item Is the concept already supported by a common virtual machine like the Java Virtual Machine or the Common Language Runtime?
 \end{enumerate}
@@
 \begin{itemize}
-  \item Managed mutation: A concurrency model may impose rules on how some state can be modified; therefore, mutation must be able to be flexibly managed.
+  \item Managed mutation and execution: A model may impose rules on how objects can be modified; therefore, mutation must be handled in a way that allows those rules to be enforced.
+Similarly, a model might restrict the invocation of operations on objects, and those rules must be enforced.
+For example, a thread should not be permitted to mutate a variable that a model declares immutable.
 
-  \item Managed execution: Similarly, invocation of operations on objects is usually restricted and must be able to be flexibly managed.
-
-  \item Ownership: An owning entity regulates mutation and execution, so ownership should be supported in a way that allows for adaptable mutation and execution rules.
+  \item Ownership: All mutation an execution on an object is regulated relative to some ``owning'' entity, so ownership should be supported in a way that allows for adaptable mutation and execution rules.
 
@@
   \item Enforceability: All defined restrictions should be enforceable on different concurrency models.
-Additionally, the semantics of the language in which a model is defined must be enforced regardless of where the model is used.
+Additionally, the semantics of the language (immutability, for example) in which a model is defined must be enforced regardless of where the model is used.
 \end{itemize}
@@
   \centering
-  \resizebox{0.63\linewidth}{!}{\input{../figures/obmop.tikz}}
+  \resizebox{0.55\linewidth}{!}{\input{../figures/obmop.tikz}}
   \caption{Marr and D'Hondt's object-based meta-object protocol~\cite{Marr2012}.}
@@
 Given those requirements, Marr and D'Hondt defined an ownership-based meta-object protocol (MOP) that could describe a low-level abstraction, illustrated in Figure~\ref{figure:obmop}.
-The owner of an object, referred to as the domain, manages operations on all of the objects that it owns (namely, reading and writing of object fields and invocation of methods on objects).
+The owner of an object, referred to as the domain, manages operations on all of the objects that it owns.
+Specifically, the domain handles all reading and writing of object fields and invocation of methods on objects.
 This satisfies the ownership requirement.
-Read and write operations are directed to the \code{readField} and \code{writeField} operations on the domain, satisfying the managed mutation requirement.
+Read and write operations are directed to the \code{read\_field} and \code{write\_field} operations on the domain, satisfying the managed mutation requirement.
 Similarly, all method invocations are directed to the \code{exec}, satisfying the managed execution requirement.
-Threads are executed in a domain and specify whether execution occurs at the language level (with restricted reflection) or the meta level (with unrestricted reflection), satisfying the leveled reflection requirement.
-Globally shared resources are handled by the domain (specifically, the \code{readGlobal} and \code{writeGlobal} methods) if they might break semantics.
+Threads are executed in a domain and specify whether execution occurs at the language level with restricted reflection or the meta level with unrestricted reflection, satisfying the leveled reflection requirement.
+Globally shared resources are handled by the domain (through the \code{read\_global} and \code{write\_global} methods) if they might break semantics.
 Additionally, the \code{primitive*} operations allow callers to override the semantics of virtual machine primitives, completing the enforcability requirement.

### 2016-12-01 paper/paper.tex
Made some final tweaks

@@
 
-Processes themselves are often composed of multiple concurrent threads of execution.
+Processes themselves are often composed of multiple concurrent threads of execution that each work on a distinct task.
 A processor can only execute one thread at a time, so multitasking is accomplished by rapidly switching between threads~\cite{Liu1973}.
@@
 The dotted line indicates the instant at which the operation takes effect.
-The lower thread invokes a read while the value of $x$ is set to $2$.
-The upper thread invokes and completes a write between the time that the read is invoked and the value is actually read.
+Thread $B$ invokes a read while the value of $x$ is set to $2$.
+Thread $A$ invokes and completes a write between the time that the read is invoked and the value is actually read.
 The read operation then returns $3$ even though $x$ was equal to $2$ at the time it was invoked.
@@
 Figure~\ref{figure:serializability} illustrates a set of serializable read and write operations.
-Even though the executions of each operation occur out of order and even intersect, the events themselves can be arranged serially.
+Even though the executions of each operation occur out of order and even intersect, the events themselves can be arranged serially; no side effects are introduced by the fact that they are executed concurrently.
 
@@
 For example, the atomic operation \code{compare-and-swap} compares the current value of a variable to a given value and only performs a write if the values are equal~\cite{Swalens2014}.
-\code{compare-and-swap} ensures that a new value cannot be written based on outdated information: Suppose that a thread $A$ reads a variable $x$ and begins computing a new value.
+\code{compare-and-swap} ensures that a new value based on outdated information cannot be written: Suppose that a thread $A$ reads a variable $x$ and begins computing a new value.
 At the same time, another thread $B$ modifies $x$.
-When $A$ tries to set the value of $x$ via \code{compare-and-swap}, the write will fail.
-As a result, all writes on $x$ are guaranteed to be linearizable.
+When $A$ tries to set the value of $x$ via \code{compare-and-swap}, the write will be denied.
 
@@
 Atomic variables often serve as the basis for higher-level concurrency control mechanisms such as semaphores and locks, which in turn support the implementation of concurrent data structures.
-A counting semaphore is an atomic variable that serves as a record of how many units of a shared resource, such as a connection or a variable, are available.
+A counting semaphore is an atomic variable that serves as a record of how many units of a shared resource, such as a connection or a pooled thread, are available.
 By taking advantage of atomic operations, the semaphore can be safely updated and will reliably determine whether a shared resource can be used.
@@
 Conceptually, STM simplifies concurrency because it allows a transaction to be thought of as a single strictly serializable operation.
-A thread cannot observe changes made to other threads while a transaction is in progress, nor can other threads observe modifications by that thread until the transaction is completed.
-Only when a transaction completes successfully will changes will become visible to other threads.
+A thread cannot observe changes made to other threads while a transaction is in progress, nor can other threads observe modifications by that thread until the transaction has completed successfully.
 Wrapping a set of operations within a transaction eliminates the need to secure locks, leading to more simple programs.
-Practically, STM is useful in any situation in which a shared object needs to be accessed and modified by multiple threads~\cite{Swalens2014}.
+A transaction that modifies several shared variables, for example, would not have to secure locks for all of them.
+Practically, STM is useful in any situation in which multiple shared objects need to be accessed and modified by multiple threads~\cite{Swalens2014}.
 
@@
 Threads communicate strictly by message passing, which avoids issues such as race conditions.
-Sending messages and reacting to received messages are inherently asynchronous processes, so communicating threads have long been a solution to describing concurrent operations.
+Because of the higher level of isolation, communicating threads have long been a solution to describing concurrent operations.
 Limiting thread communication to message passing also enforces modularity and limits possible negative side effects.
@@
 Other models, most notably the actor model, rely on asynchronous message passing to specific entities (actors) rather than named channels~\cite{Agha1986}.
-Actors can only communicate by asynchronous message passing.
 Upon receiving a message, an actor can choose to perform an operation on a resource that it controls, send messages to other actors, create new actors, determine the behavior used for the next received message, or ignore the message entirely.
@@
 Proxies are a general term used to describe placeholders for values that are the result of some concurrently executed operation.
-Unlike communicating thread models, which rely on message passing between dedicated threads, a proxy executes some function in a new thread and delivers the result upon completion~\cite{Swalens2014}.
+Unlike communicating thread models, which rely on message passing between dedicated threads, a proxy executes some task in a new thread and delivers the result upon completion~\cite{Swalens2014}.
 
@@
 In practice, proxies are used to execute long-running operations such as rendering or network requests~\cite{Swalens2014} without blocking other operations.
-Promise objects, which can be easily passed and composed, offer more flexibility than constructs such as asynchronous callbacks.
+For example, to eliminate the need for a program to block on a long network request, a promise could be used to complete the network request in another thread and specify an action that should take place once the result becomes available.
+Promise objects, which can be easily passed, composed, and chained, offer more flexibility than constructs such as asynchronous callbacks.
+In fact, many mainstream languages now directly support proxies.
 
@@
 
-Even if concurrency models guarantee their own correctness, issues may arise when different models are used together.
+Even if a concurrency model guarantees correctness when used alone, issues may arise when it is used with other models.
 For example, an implementation of an STM that guarantees linearizability assumes that all shared resources are managed by the STM~\cite{Shavit1995}.
@@
 Swalens et al.~\cite{Swalens2014} attempted to identify issues that arise when different concurrency models are combined, hoping that future work could identify common ``building blocks'' that could be used to compose a variety of concurrency models.
-The study surveyed all of the concurrency models available in the Clojure programming language and used them in within one another to see what types of correctness issues were encountered.
+The study surveyed all of the concurrency models available in the Clojure programming language and used them within one another to see what types of correctness issues were encountered.
 
@@
 This property is also referred to as partial correctness.
-Safety is generally achieved in the management of shared resources.
+In the context of concurrent programming, safety is generally achieved by the management of shared resources.
 STM, for instance, only allows shared memory to be accessed through transactions; communicating threads only allow data to be shared through message passing.
@@
   \item A model might reexecute code containing another concurrency model that performs an irrevocable action.
-For instance, if a transaction were to contain an operation that sent a message on a channel, reexecutions of the transaction would cause the message to be sent repeatedly.
+If a block of code such as a transaction were to contain an operation that sent a message on a channel, reexecutions of that block would cause the message to be sent repeatedly.
 
@@
 
-  \item A model might not guarantee safety or liveness by design, making composability impossible.
-Perhaps obviously, a model that does not guarantee safety or liveness would violate the safety guarantee of an outer model unless such a conflict was explicitly prevented.
+  \item A model might not guarantee safety or liveness by design, making safe or live composability impossible.
 \end{itemize}
 
-The study noted that some types of bad interactions are prevented by Clojure; for instance, sending a message to an agent from within a transaction is delayed until the transaction succeeds.
+The study noted that some types of bad interactions are prevented by Clojure; as an example, sending a message to an agent from within a transaction is delayed until the transaction succeeds.
 The study also noted that similar safeguards could prevent other negative interactions.
@@
 A unifying abstraction would enable each concurrency model in a high-level language to be represented in machine code without sacrificing performance or semantics.
-For instance, a concurrency model in Clojure could be accurately represented by an underlying abstraction in the Java Virtual Machine, and code written in the Java language could interact with that model correctly.
+For instance, a concurrency model in Clojure could be accurately represented by an underlying abstraction in the Java Virtual Machine, and code written in the Java programming language could interact with that model correctly.
 To identify the concepts that would need to be supported at the virtual machine level, Marr and D'Hondt selected four questions:
@@
 Semantic guarantees may be enforced by a compiler but not the virtual machine.
-For instance, immutable objects in one language may not be immutable in machine code, and therefore may be mutated by another language.
+For example, immutable objects in one language may not be immutable in machine code, and therefore may be mutated by another language.
 
@@
 
-From those questions, 26 concepts were identified that would benefit from better low-level support.
-Of those, 18 suffered from loss of semantic guarantees when compiled to machine code.
-With that in mind, Marr and D'Hondt derived the following requirements for a unifying abstraction:
+From those questions, 26 concepts were identified that would benefit from runtime support.
+Of those, 18 suffered from loss of semantic guarantees when compiled to machine code; the survey mentioned transactions as an example.
+With those concepts in mind, Marr and D'Hondt derived the following requirements for a unifying abstraction:
 
@@
   \item Managed mutation and execution: A model may impose rules on how objects can be modified; therefore, mutation must be handled in a way that allows those rules to be enforced.
-Similarly, a model might restrict the invocation of operations on objects, and those rules must be enforced.
-For example, a thread should not be permitted to mutate a variable that a model declares immutable.
+By way of example, a thread should not be permitted to mutate a variable that a model declares immutable.
+Similarly, a model may restrict the invocation of operations on objects.
 
-  \item Ownership: All mutation an execution on an object is regulated relative to some ``owning'' entity, so ownership should be supported in a way that allows for adaptable mutation and execution rules.
+  \item Ownership: All mutation and execution on an object is regulated relative to some ``owning'' entity, so ownership should be supported in a way that allows for adaptable mutation and execution rules.
 
@@
   \resizebox{0.55\linewidth}{!}{\input{../figures/obmop.tikz}}
-  \caption{Marr and D'Hondt's object-based meta-object protocol~\cite{Marr2012}.}
+  \caption{Marr and D'Hondt's owenership-based meta-object protocol~\cite{Marr2012}.}
 \label{figure:obmop}
@@
 The MOP was able to support LRSTM (a STM for the Smalltalk language), Clojure's agents (which rely on communicating threads), active objects (a variation on proxies), and several other models.
-While the MOP was able to successfully enforce the semantics of each model, the resulting performance cost suggested that the MOP must be implemented at the virtual machine level to sufficiently reduce the abstraction overhead.
+While the MOP was able to successfully enforce the semantics of each model, the resulting performance cost suggested that the MOP must be implemented at the virtual machine level to sufficiently reduce the overhead of abstraction.
 Marr and D'Hondt also note that regarding the owner of an object as the only entity able to restrict interaction is somewhat limiting.
